---
# title: "ATRAPA Work Package 1: Built Environment Transformations Dataset"
format: html
editor: visual
bibliography: wp1.bib
---

## OSM validation method

### Aim

Validate OSM cycling infrastructure in Barcelona across years 2019 and 2023.

### Design

-   Stratified random sample of census tracts (centrality × density, 3×3).

-   Five tracts per stratum (45 in total).

-   Per tract and year:

    -   3 random points on the OSM cycleway network (*if present in tract*) → check for false positives for that year.

    -   3 random points on the OSM general road network → check for false negative for that year.

-   Validation source: Google Street View (GSV).

### OSM data

```{r include=FALSE}
#remotes::install_github("ropensci/osmextract")
library(osmdata)
library(osmextract)
library(sf)
library(tidyverse)
library(tmap)
library(grid)
library(gridExtra)
library(spdep)
```

```{r boundaries}
#| label: boundaries
#| eval: false
#| include: false

bb <- getbb("Barcelona, Spain")

od <- opq(bb) |>
  add_osm_feature("boundary", "administrative") |>
  add_osm_feature("admin_level", "8") |>
  add_osm_feature("name", "Barcelona") |>
  osmdata_sf()

perim <- od$osm_multipolygons
if (is.null(perim) || nrow(perim) == 0) perim <- od$osm_polygons
stopifnot(nrow(perim) > 0)

# Prefer municipality (Wikidata Q1492) else largest polygon
if ("wikidata" %in% names(perim) && any(perim$wikidata == "Q1492")) {
  perim <- perim[perim$wikidata == "Q1492", ]
} else {
  perim <- perim[which.max(st_area(perim)), ]
}

perim <- st_make_valid(perim) |> st_cast("MULTIPOLYGON") |> st_set_crs(4326)

st_write(perim, "data/barcelona_perimeter.gpkg",
         layer = "barcelona_perimeter", driver = "GPKG", append = FALSE)
```

```{r infras}
#| label: infras
#| eval: false
#| include: false
perimeter_path <- "data/barcelona_perimeter.gpkg"
stopifnot(file.exists(perimeter_path))
city_perimeter <- sf::st_read(perimeter_path, layer = "barcelona_perimeter", quiet = TRUE)
if (is.na(sf::st_crs(city_perimeter)$epsg) || sf::st_crs(city_perimeter)$epsg != 4326) {
  city_perimeter <- sf::st_transform(city_perimeter, 4326)
}

allowed_highway <- c(
  "living_street","pedestrian","cycleway","motorway","trunk","primary","secondary",
  "tertiary","unclassified","residential","motorway_link","trunk_link","primary_link",
  "secondary_link","tertiary_link","service","track","bus_guideway","escape","raceway","busway"
)

fetch_and_crop_data <- function(version, perimeter, output_file) {
  if (file.exists(output_file)) {
    message("↪ Skipping (exists): ", basename(output_file)); return(invisible(NULL))
  }
  # Always pull from "spain" (works for 2017/2019/2023), read only 'lines'
  lines_data <- tryCatch({
    oe_get(
      place   = "spain",
      version = version,
      layer   = "lines",
      boundary = st_bbox(perimeter), boundary_type = "clipsrc",  # fast pre-clip
      extra_tags = c("cycleway","cycleway:left","cycleway:right","bicycle","route"),
      quiet = FALSE
    )
  }, error = function(e) { message("Error fetching ", version, ": ", e$message); return(NULL) })

  if (is.null(lines_data) || nrow(lines_data) == 0) {
    message("No data returned for version ", version); return(invisible(NULL))
  }

  # Keep only (multi)lines
  is_line <- sf::st_geometry_type(lines_data) %in% c("LINESTRING","MULTILINESTRING")
  lines_data <- lines_data[is_line, ]
  if (nrow(lines_data) == 0) { message("No LINESTRINGs for version ", version); return(invisible(NULL)) }

  # Filter to common road classes (do it in R to avoid GDAL vectortranslate issues)
  if ("highway" %in% names(lines_data)) {
    lines_data <- dplyr::filter(lines_data, highway %in% allowed_highway)
  }

  # Final polygon clip
  perim_valid <- sf::st_make_valid(perimeter)
  lines_data <- lines_data[sf::st_intersects(lines_data, perim_valid, sparse = FALSE), ]
  lines_data <- sf::st_intersection(lines_data, perim_valid)

  out_layer <- sub("\\.gpkg$", "", basename(output_file))
  sf::st_write(lines_data, output_file, layer = out_layer, driver = "GPKG", append = FALSE, quiet = TRUE)
  message("✓ Saved ", basename(output_file), " (", nrow(lines_data), " features)")
}

# Exactly these years (add 2017 if you want by uncommenting)
# fetch_and_crop_data("170101", city_perimeter, "data/barcelona_170101_lines.gpkg")
fetch_and_crop_data("200101", city_perimeter, "data/barcelona_19_lines.gpkg")
fetch_and_crop_data("240101", city_perimeter, "data/barcelona_23_lines.gpkg")

message("Done: Barcelona (files in data/)")
```

```{r}
#| label: infraplot
#| eval: false
#| include: false
# --- cities to plot (match your file prefixes under data/) ---
city_tags <- c("barcelona")      # we saved data/barcelona_*.gpkg
versions  <- c("19","23")  # 2019 & 2023

# --- palette & helpers ---
pal <- c("cycleway"="#E41A1C","pedestrian"="#377EB8","living street"="#4DAF4A","others"="grey")

read_perimeter <- function(tag){
  perim_path <- file.path("data", paste0(tag, "_perimeter.gpkg"))  # << no subfolder
  if (!file.exists(perim_path)) return(NULL)
  # read specific layer and ensure WGS84
  sf::st_read(perim_path, layer = paste0(tag, "_perimeter"), quiet = TRUE) |>
    sf::st_transform(4326)
}

read_infra <- function(tag, ver){
  gpkg <- file.path("data", paste0(tag, "_", ver, "_lines.gpkg"))  # << no subfolder
  lyr  <- paste0(tag, "_", ver, "_lines")
  if (!file.exists(gpkg)) return(NULL)
  x <- tryCatch(sf::st_read(gpkg, layer = lyr, quiet = TRUE), error = function(e) NULL)
  x
}

mk_aggr <- function(x, perimeter){
  if (is.null(x) || nrow(x) == 0) return(NULL)
  x <- sf::st_transform(x, sf::st_crs(perimeter))
  if (!"highway_aggr" %in% names(x)) {
    x <- x |>
      dplyr::mutate(highway_aggr = dplyr::case_when(
        highway == "cycleway"      ~ "cycleway",
        highway == "pedestrian"    ~ "pedestrian",
        highway == "living_street" ~ "living street",
        TRUE ~ "others"
      ))
  }
  x
}

mk_map <- function(perimeter, x, title){
  if (is.null(x) || nrow(x) == 0) {
    return(tm_shape(perimeter) + tm_borders() + tm_title(paste0(title, " (no data)")))
  }
  tm_shape(perimeter) + tm_borders() +
    tm_shape(x) +
    tm_lines(col = "highway_aggr", lwd = 0.5, col_alpha = 0.6,
             col.scale = tm_scale(values = pal)) +
    tm_title(title)
}

make_city_panel <- function(city_tag){
  perim <- read_perimeter(city_tag)
  if (is.null(perim)) return(NULL)

  x19 <- mk_aggr(read_infra(city_tag, "19"), perim)
  x23 <- mk_aggr(read_infra(city_tag, "23"), perim)

  m19 <- mk_map(perim, x19, "2019")   # << fixed var name
  m23 <- mk_map(perim, x23, "2023")   # << fixed var name
  tmap_arrange(m19, m23, nrow = 1)
}

# --- display all city panels ---
tmap_mode("view")
panels <- lapply(city_tags, make_city_panel)
invisible(lapply(panels[!sapply(panels, is.null)], print))
```

### Tract data

```{r}
#| include: false
# ================================================================
# STRATIFIED SAMPLING AND STREET VIEW LINK GENERATION
# ================================================================

# ---- Load required libraries ----
library(sf)
library(dplyr)
library(tidyr)
library(biscale)
library(ggplot2)
library(cowplot)
library(osmdata)
library(purrr)
library(jsonlite)
library(httr)
library(tmap)
library(stringr)
library(sf)
library(leaflet)
library(htmltools)
```

```{r}
#| include: false
# ================================================================
# 1. LOAD AND PREPARE DATA
# ================================================================

# Load spatial and population data
tracts_geo <- readRDS("data/bcn-seccio-censal/BCN_seccion_censal_2015.rds")
tracts_pop <- readRDS("data/pop/Population_Census_tract_2015_2022.rds")

tracts_pop_wide <- tracts_pop |> 
  filter(Year %in% c("2015", "2019", "2022")) |> 
  pivot_wider(names_from = Year, values_from = Population, names_prefix = "pop_")

# Merge geometries with population data
barcelona_tracts <- left_join(tracts_geo, tracts_pop_wide, by = c("CUSEC" = "GEOID")) |> 
  st_transform(25831)

# Compute area, population density, and distance to city center
placa_catalunya <- st_sfc(st_point(c(431870, 4581450)), crs = 25831)

barcelona_tracts <- barcelona_tracts |> 
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    dens_2015 = pop_2015 / area_km2,
    dens_2019 = pop_2019 / area_km2,
    dens_2022 = pop_2022 / area_km2,
    centroid = st_centroid(geometry),
    dist_centre_km = as.numeric(st_distance(centroid, placa_catalunya)) / 1000
  )
```

```{r}
#| message: false
#| warning: false
#| include: false
#| paged-print: false

# ================================================================
# 2. STRATIFIED SAMPLING
# ================================================================

barcelona_tracts <- barcelona_tracts |>
  mutate(
    dens_stratum = ntile(dens_2022, 3),
    cent_stratum = ntile(-dist_centre_km, 3),
    stratum_id = paste0("D", dens_stratum, "_C", cent_stratum)
  )

set.seed(123)
per_stratum <- 5  # try 3–6 depending on effort

sampled_tracts <- barcelona_tracts |> 
  group_by(stratum_id) |> 
  sample_n(size = min(per_stratum, n())) |> 
  ungroup()
```

```{r}
#| echo: false

# ================================================================
# 3. BIVARIATE MAPPING
# ================================================================

barcelona_tracts <- barcelona_tracts |> 
mutate(centrality_flipped = -dist_centre_km)
 
bb_bivar <- bi_class(barcelona_tracts, x = dens_2022, y = centrality_flipped, style = "quantile", dim = 3)
 
bivar_map <- ggplot() +
   geom_sf(data = bb_bivar, aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
   bi_scale_fill(pal = "DkBlue2", dim = 3) +
   geom_sf(data = sampled_tracts, fill = NA, color = "black", linewidth = 0.5) +
   labs(title = "") +
   bi_theme() +
   theme(plot.title = element_text(size = 12, hjust = 0.5))
 
 bivar_legend <- bi_legend(pal = "DkBlue2", dim = 3, xlab = "Higher Density", ylab = "More Central", size = 9) +
   theme(axis.title = element_text(size = 9), axis.text = element_blank(),
         plot.margin = margin(5, 5, 5, 5), panel.background = element_blank(), plot.background = element_blank())
 
 final_plot <- ggdraw() +
   draw_plot(bivar_map, 0, 0, 1, 1) +
   draw_plot(bivar_legend, x = 0.73, y = 0.03, width = 0.26, height = 0.26)
 
 final_plot
 
 ggsave("figs/stratified_sample_bivariate_map.png", final_plot, width = 8, height = 8, dpi = 300)
```

### GSV data

```{r}
#| label: setup-io
#| message: false
#| warning: false
#| include: false

library(sf); library(dplyr); library(stringr); library(leaflet)

# ---- params ----
city_tag <- "barcelona"
ver19    <- "19"
ver23    <- "23"
crs_work <- 25831
tol_m    <- 10
min_len  <- 30

# ---- paths ----
dir_city <- "data"
gpkg19   <- file.path(dir_city, paste0(city_tag, "_", ver19, "_lines.gpkg"))
gpkg23   <- file.path(dir_city, paste0(city_tag, "_", ver23, "_lines.gpkg"))
lyr19    <- paste0(city_tag, "_", ver19, "_lines")
lyr23    <- paste0(city_tag, "_", ver23, "_lines")

stopifnot(file.exists(gpkg19), file.exists(gpkg23))

# ---- read ----
l19 <- st_read(gpkg19, layer = lyr19, quiet = TRUE)
l23 <- st_read(gpkg23, layer = lyr23, quiet = TRUE)
```

```{r}
#| label: cycle-change-helpers
#| message: false
#| warning: false
#| include: false

library(sf); library(dplyr); library(stringr)

# 1) STRICT cycleway picker — only clear, dedicated cycling infrastructure
pick_cycle_strict <- function(x) {
  stopifnot(inherits(x, "sf"))
  if (nrow(x) == 0) return(x)
  
  n  <- nrow(x); nm <- names(x)
  get_chr <- function(col) if (col %in% nm) as.character(x[[col]]) else rep(NA_character_, n)
  get_lwr <- function(col) tolower(get_chr(col))
  
  highway      <- get_lwr("highway")
  bicycle      <- get_lwr("bicycle")
  bicycle_road <- get_lwr("bicycle_road")
  
  # 1) DEDICATED CYCLEWAYS (highest priority)
  is_cycleway <- highway == "cycleway"
  
  # 2) CYCLE LANES/TRACKS on roads (must be physical infrastructure)
  lane_cols <- c("cycleway", "cycleway:both", "cycleway:left", "cycleway:right")
  lane_vals <- c("lane", "track", "opposite_lane", "opposite_track", "separate")
  
  has_physical_lane <- rep(FALSE, n)
  for (col in lane_cols) {
    if (col %in% nm) {
      vals <- get_lwr(col)
      has_physical_lane <- has_physical_lane | (vals %in% lane_vals)
    }
  }
  
  # 3) DESIGNATED cycling paths (must be explicitly designated for bicycles)
  is_designated_path <- (highway %in% c("path", "footway", "track")) & 
                        (bicycle %in% c("designated", "yes"))
  
  # 4) BICYCLE ROADS
  is_bicycle_road <- bicycle_road == "yes"
  
  # 5) EXCLUDE these common false positives:
  # - Shared lanes (sharrows)
  has_sharrow <- rep(FALSE, n)
  for (col in lane_cols) {
    if (col %in% nm) {
      vals <- get_lwr(col)
      has_sharrow <- has_sharrow | (vals %in% c("shared_lane", "sharrow"))
    }
  }
  
  # - Permmissive access only (bicycle=yes but no physical infra)
  is_permissive_only <- (bicycle == "yes") & 
                        !is_cycleway & 
                        !has_physical_lane & 
                        !is_designated_path
  
  # - Footways/paths where bicycles are merely tolerated
  is_tolerated_path <- (highway %in% c("footway", "path")) & 
                       !(bicycle %in% c("designated", "yes"))
  
  # COMBINE: Keep only clear cycling infrastructure, exclude ambiguous cases
  keep <- (is_cycleway | has_physical_lane | is_designated_path | is_bicycle_road) &
          !has_sharrow & 
          !is_permissive_only & 
          !is_tolerated_path
  
  # Additional safety: exclude major roads without explicit cycling infrastructure
  is_major_road <- highway %in% c("motorway", "motorway_link", "trunk", "trunk_link",
                                 "primary", "primary_link", "secondary", "secondary_link")
  keep <- keep & !(is_major_road & !has_physical_lane)
  
  result <- x[which(keep), , drop = FALSE]
  
  # Debug output (optional)
  cat("Strict filtering: kept", nrow(result), "of", n, "segments\n")
  
  return(result)
}

# [Rest of your functions remain the same...]
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
# --- Minimal change mapping: 2019 vs 2023 (cycling) --------------------------
library(dplyr)
library(stringr)
library(sf)

# --- Revised params ---
crs_work <- 25831      # Barcelona UTM; must be a metric CRS
tol_m    <- 10          # REDUCED from 10m to 2m for cycling infrastructure
min_len  <- 30         # drop slivers shorter than this (meters)

# Now use the functions
cyc19 <- l19 |> pick_cycle_strict() |> st_make_valid()
cyc23 <- l23 |> pick_cycle_strict() |> st_make_valid()

# Ensure both are in the same CRS
cyc19 <- st_transform(cyc19, crs_work)
cyc23 <- st_transform(cyc23, crs_work)

# ==== REVISED 3) Added/Removed with improved geometry handling ====

# 3.1 Improved geometry normalization with better error handling
normalize_geoms <- function(x) {
  x <- st_make_valid(x)
  x <- st_cast(x, "MULTILINESTRING", warn = FALSE)
  x <- st_line_merge(x)
  # Extract individual linestrings more carefully
  geoms <- st_geometry(x)
  individual_lines <- st_cast(geoms, "LINESTRING", warn = FALSE)
  st_set_geometry(x, individual_lines) |> st_make_valid()
}

cyc19_n <- normalize_geoms(cyc19)
cyc23_n <- normalize_geoms(cyc23)

# 3.2 Build buffers with smaller tolerance
buf19 <- st_buffer(st_geometry(cyc19_n), tol_m) |> st_union()
buf23 <- st_buffer(st_geometry(cyc23_n), tol_m) |> st_union()

# 3.3 Calculate differences with improved handling
added_raw <- st_difference(st_geometry(cyc23_n), buf19)
removed_raw <- st_difference(st_geometry(cyc19_n), buf23)

# 3.4 Process results with better geometry extraction
process_difference <- function(diff_geom, crs) {
  if (length(diff_geom) == 0) {
    return(st_sf(geometry = st_sfc(), crs = crs))
  }
  
  # Handle geometry collections properly
  diff_geom <- st_collection_extract(diff_geom, "LINESTRING")
  diff_geom <- st_cast(diff_geom, "LINESTRING", warn = FALSE)
  
  # Remove zero-length geometries
  lengths <- st_length(diff_geom)
  diff_geom <- diff_geom[as.numeric(lengths) > 0.1]  # > 10cm
  
  st_sf(geometry = diff_geom, crs = crs)
}

added_partial <- process_difference(added_raw, crs_work)
removed_partial <- process_difference(removed_raw, crs_work)

# 3.5 Length filtering function (single definition)
len_ok <- function(x, min_len = 10){
  if (!inherits(x, "sf") || nrow(x) == 0) return(x)
  len_m <- as.numeric(st_length(x))
  x <- x[len_m >= min_len, ]
  x
}

added <- len_ok(added_partial, min_len = min_len)
removed <- len_ok(removed_partial, min_len = min_len)

# ==== 4) Debugging: Check what's happening ====
cat("2019 segments:", nrow(cyc19_n), "\n")
cat("2023 segments:", nrow(cyc23_n), "\n")
cat("Added segments:", nrow(added), "\n")
cat("Removed segments:", nrow(removed), "\n")

# Check if specific 2023 segments are missing
if (nrow(added) == 0 & nrow(cyc23_n) > 0) {
  cat("Warning: No segments detected as added. Possible issues:\n")
  cat("1. Tolerance too large (current:", tol_m, "m)\n")
  cat("2. CRS mismatch\n")
  cat("3. Geometry issues\n")
  
  # Test: check if buffers are covering everything
  coverage_test <- st_covered_by(st_geometry(cyc23_n), buf19)
  cat("2023 segments fully covered by 2019 buffer:", sum(coverage_test), "/", nrow(cyc23_n), "\n")
}

# ==== 5) Midpoints for quick validation sampling ====
midpoint_one <- function(x) {
  # one sample per line: midpoint
  pts <- st_line_sample(x, sample = 0.5)
  st_as_sf(data.frame(id = seq_along(pts)), geometry = st_cast(pts, "POINT"), crs = crs_work)
}

added_pts   <- midpoint_one(added)   |> mutate(class = "ADD")
removed_pts <- midpoint_one(removed) |> mutate(class = "REMOVE")

# ==== 6) Quick map ====
leaflet() |>
  addProviderTiles("CartoDB.Positron") |>
  addPolylines(data = st_transform(cyc19_n, 4326), group = "2019",
               weight = 3, color = "#377EB8") |>
  addPolylines(data = st_transform(cyc23_n, 4326), group = "2023",
               weight = 3, color = "#4DAF4A") |>
  addPolylines(data = st_transform(removed, 4326), group = "Removed",
               weight = 3, color = "white") |>
  addPolylines(data = st_transform(added, 4326), group = "Added",
               weight = 3, color = "purple") |>
  addLayersControl(
    overlayGroups = c("2019","2023","Removed","Added"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  addLegend(
    position = "bottomright",
    title = "Layers",
    colors = c("#377EB8", "#4DAF4A", "white", "purple"),
    labels = c("2019", "2023", "Removed", "Added"),
    opacity = 1
  )

  
```

```{r}
# Excel to check the points
```


<!-- ## References -->
