---
# title: "ATRAPA Work Package 1: Built Environment Transformations Dataset"
format: html
editor: visual
bibliography: wp1.bib
---

# Method to Validate OSM Cycling Transformations (2019→2023)

## Introduction

- Context: Urban transformations, such as new bike lanes and pedestrian areas, can strongly influence mobility behaviour and health. Studying these effects requires reliable historical data on when and where such changes occurred.

- Problem: Official, standardised datasets of infrastructure transformations are rarely available across cities and years. This gap makes it difficult to analyse behavioural impacts.

- Aim: quantify how well OSM detects cycling infrastructure (CI) changes (adds/removals) and estimate missed additions.

- Outputs: precision for ADD/REMOVE, false-negative (FN) rate for additions, recall for additions, calibrated (error-adjusted) change estimates with CIs, by-stratum diagnostics.

## Data and Methods 

### Setting and period

- Study area: Barcelona; period: 2019→2023 (snapshot dates: 2020-01-01 for 2019 and 2024-01-01 for 2023; specify exact extracts).

### Network data (OSM)

- Extraction. Download OSM linework for both years.

- CI selector. Keep: highway=cycleway, bicycle-designated path/track, roads with cycleway:* tags, bicycle_road=yes, designated living_street/pedestrian.

- Non-CI (general network). Base roads excluding all CI tags (strict complement).

- Change detection. Build ADDED (present in 2023, absent 2019) and REMOVED (present 2019, absent 2023) layers.

```{r include=FALSE}
#remotes::install_github("ropensci/osmextract")
library(osmdata)
library(osmextract)
library(sf)
library(tidyverse)
library(tmap)
library(grid)
library(gridExtra)
library(spdep)
```

```{r}
#| label: boundaries
#| eval: false
#| include: false

bb <- getbb("Barcelona, Spain")

od <- opq(bb) |>
  add_osm_feature("boundary", "administrative") |>
  add_osm_feature("admin_level", "8") |>
  add_osm_feature("name", "Barcelona") |>
  osmdata_sf()

perim <- od$osm_multipolygons
if (is.null(perim) || nrow(perim) == 0) perim <- od$osm_polygons
stopifnot(nrow(perim) > 0)

# Prefer municipality (Wikidata Q1492) else largest polygon
if ("wikidata" %in% names(perim) && any(perim$wikidata == "Q1492")) {
  perim <- perim[perim$wikidata == "Q1492", ]
} else {
  perim <- perim[which.max(st_area(perim)), ]
}

perim <- st_make_valid(perim) |> st_cast("MULTIPOLYGON") |> st_set_crs(4326)

st_write(perim, "data/barcelona_perimeter.gpkg",
         layer = "barcelona_perimeter", driver = "GPKG", append = FALSE)
```

```{r}
#| label: infras
#| eval: false
#| include: false
perimeter_path <- "data/barcelona_perimeter.gpkg"
stopifnot(file.exists(perimeter_path))
city_perimeter <- sf::st_read(perimeter_path, layer = "barcelona_perimeter", quiet = TRUE)
if (is.na(sf::st_crs(city_perimeter)$epsg) || sf::st_crs(city_perimeter)$epsg != 4326) {
  city_perimeter <- sf::st_transform(city_perimeter, 4326)
}

allowed_highway <- c(
  "living_street","pedestrian","cycleway","motorway","trunk","primary","secondary",
  "tertiary","unclassified","residential","motorway_link","trunk_link","primary_link",
  "secondary_link","tertiary_link","service","track","bus_guideway","escape","raceway","busway"
)

fetch_and_crop_data <- function(version, perimeter, output_file) {
  if (file.exists(output_file)) {
    message("↪ Skipping (exists): ", basename(output_file)); return(invisible(NULL))
  }
  # Always pull from "spain" (works for 2017/2019/2023), read only 'lines'
  lines_data <- tryCatch({
    oe_get(
      place   = "spain",
      version = version,
      layer   = "lines",
      boundary = st_bbox(perimeter), boundary_type = "clipsrc",  # fast pre-clip
      extra_tags = c("cycleway","cycleway:left","cycleway:right","bicycle","route"),
      quiet = FALSE
    )
  }, error = function(e) { message("Error fetching ", version, ": ", e$message); return(NULL) })

  if (is.null(lines_data) || nrow(lines_data) == 0) {
    message("No data returned for version ", version); return(invisible(NULL))
  }

  # Keep only (multi)lines
  is_line <- sf::st_geometry_type(lines_data) %in% c("LINESTRING","MULTILINESTRING")
  lines_data <- lines_data[is_line, ]
  if (nrow(lines_data) == 0) { message("No LINESTRINGs for version ", version); return(invisible(NULL)) }

  # Filter to common road classes (do it in R to avoid GDAL vectortranslate issues)
  if ("highway" %in% names(lines_data)) {
    lines_data <- dplyr::filter(lines_data, highway %in% allowed_highway)
  }

  # Final polygon clip
  perim_valid <- sf::st_make_valid(perimeter)
  lines_data <- lines_data[sf::st_intersects(lines_data, perim_valid, sparse = FALSE), ]
  lines_data <- sf::st_intersection(lines_data, perim_valid)

  out_layer <- sub("\\.gpkg$", "", basename(output_file))
  sf::st_write(lines_data, output_file, layer = out_layer, driver = "GPKG", append = FALSE, quiet = TRUE)
  message("✓ Saved ", basename(output_file), " (", nrow(lines_data), " features)")
}

# Exactly these years (add 2017 if you want by uncommenting)
# fetch_and_crop_data("170101", city_perimeter, "data/barcelona_170101_lines.gpkg")
fetch_and_crop_data("200101", city_perimeter, "data/barcelona_19_lines.gpkg")
fetch_and_crop_data("240101", city_perimeter, "data/barcelona_23_lines.gpkg")

message("Done: Barcelona (files in data/)")
```

```{r}
#| label: setup-io
#| message: false
#| warning: false
#| include: false

library(sf); library(dplyr); library(stringr); library(leaflet)

# ---- params ----
city_tag <- "barcelona"
ver19    <- "19"
ver23    <- "23"
crs_work <- 25831
tol_m    <- 10
min_len  <- 30

# ---- paths ----
dir_city <- "data"
gpkg19   <- file.path(dir_city, paste0(city_tag, "_", ver19, "_lines.gpkg"))
gpkg23   <- file.path(dir_city, paste0(city_tag, "_", ver23, "_lines.gpkg"))
lyr19    <- paste0(city_tag, "_", ver19, "_lines")
lyr23    <- paste0(city_tag, "_", ver23, "_lines")

stopifnot(file.exists(gpkg19), file.exists(gpkg23))

# ---- read ----
l19 <- st_read(gpkg19, layer = lyr19, quiet = TRUE)
l23 <- st_read(gpkg23, layer = lyr23, quiet = TRUE)
```

```{r}
#| label: cycle-change-helpers
#| message: false
#| warning: false
#| include: false

library(sf); library(dplyr); library(stringr)

# 1) STRICT cycleway picker — only clear, dedicated cycling infrastructure
pick_cycle_strict <- function(x) {
  stopifnot(inherits(x, "sf"))
  if (nrow(x) == 0) return(x)
  
  n  <- nrow(x); nm <- names(x)
  get_chr <- function(col) if (col %in% nm) as.character(x[[col]]) else rep(NA_character_, n)
  get_lwr <- function(col) tolower(get_chr(col))
  
  highway      <- get_lwr("highway")
  bicycle      <- get_lwr("bicycle")
  bicycle_road <- get_lwr("bicycle_road")
  
  # 1) DEDICATED CYCLEWAYS (highest priority)
  is_cycleway <- highway == "cycleway"
  
  # 2) CYCLE LANES/TRACKS on roads (must be physical infrastructure)
  lane_cols <- c("cycleway", "cycleway:both", "cycleway:left", "cycleway:right")
  lane_vals <- c("lane", "track", "opposite_lane", "opposite_track", "separate")
  
  has_physical_lane <- rep(FALSE, n)
  for (col in lane_cols) {
    if (col %in% nm) {
      vals <- get_lwr(col)
      has_physical_lane <- has_physical_lane | (vals %in% lane_vals)
    }
  }
  
  # 3) DESIGNATED cycling paths (must be explicitly designated for bicycles)
  is_designated_path <- (highway %in% c("path", "footway", "track")) & 
                        (bicycle %in% c("designated", "yes"))
  
  # 4) BICYCLE ROADS
  is_bicycle_road <- bicycle_road == "yes"
  
  # 5) EXCLUDE these common false positives:
  # - Shared lanes (sharrows)
  has_sharrow <- rep(FALSE, n)
  for (col in lane_cols) {
    if (col %in% nm) {
      vals <- get_lwr(col)
      has_sharrow <- has_sharrow | (vals %in% c("shared_lane", "sharrow"))
    }
  }
  
  # - Permmissive access only (bicycle=yes but no physical infra)
  is_permissive_only <- (bicycle == "yes") & 
                        !is_cycleway & 
                        !has_physical_lane & 
                        !is_designated_path
  
  # - Footways/paths where bicycles are merely tolerated
  is_tolerated_path <- (highway %in% c("footway", "path")) & 
                       !(bicycle %in% c("designated", "yes"))
  
  # COMBINE: Keep only clear cycling infrastructure, exclude ambiguous cases
  keep <- (is_cycleway | has_physical_lane | is_designated_path | is_bicycle_road) &
          !has_sharrow & 
          !is_permissive_only & 
          !is_tolerated_path
  
  # Additional safety: exclude major roads without explicit cycling infrastructure
  is_major_road <- highway %in% c("motorway", "motorway_link", "trunk", "trunk_link",
                                 "primary", "primary_link", "secondary", "secondary_link")
  keep <- keep & !(is_major_road & !has_physical_lane)
  
  result <- x[which(keep), , drop = FALSE]
  
  return(result)
}

# [Rest of your functions remain the same...]
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| paged-print: false
# --- Minimal change mapping: 2019 vs 2023 (cycling) --------------------------
library(dplyr)
library(stringr)
library(sf)

# --- Revised params ---
crs_work <- 25831      # Barcelona UTM; must be a metric CRS
tol_m    <- 10          # REDUCED from 10m to 2m for cycling infrastructure
min_len  <- 30         # drop slivers shorter than this (meters)

# Now use the functions
cyc19 <- l19 |> pick_cycle_strict() |> st_make_valid()
cyc23 <- l23 |> pick_cycle_strict() |> st_make_valid()

# Ensure both are in the same CRS
cyc19 <- st_transform(cyc19, crs_work)
cyc23 <- st_transform(cyc23, crs_work)

# ==== REVISED 3) Added/Removed with improved geometry handling ====

# 3.1 Improved geometry normalization with better error handling
normalize_geoms <- function(x) {
  x <- st_make_valid(x)
  x <- st_cast(x, "MULTILINESTRING", warn = FALSE)
  x <- st_line_merge(x)
  # Extract individual linestrings more carefully
  geoms <- st_geometry(x)
  individual_lines <- st_cast(geoms, "LINESTRING", warn = FALSE)
  st_set_geometry(x, individual_lines) |> st_make_valid()
}

cyc19_n <- normalize_geoms(cyc19)
cyc23_n <- normalize_geoms(cyc23)

# 3.2 Build buffers with smaller tolerance
buf19 <- st_buffer(st_geometry(cyc19_n), tol_m) |> st_union()
buf23 <- st_buffer(st_geometry(cyc23_n), tol_m) |> st_union()

# 3.3 Calculate differences with improved handling
added_raw <- st_difference(st_geometry(cyc23_n), buf19)
removed_raw <- st_difference(st_geometry(cyc19_n), buf23)

# 3.4 Process results with better geometry extraction
process_difference <- function(diff_geom, crs) {
  if (length(diff_geom) == 0) {
    return(st_sf(geometry = st_sfc(), crs = crs))
  }
  
  # Handle geometry collections properly
  diff_geom <- st_collection_extract(diff_geom, "LINESTRING")
  diff_geom <- st_cast(diff_geom, "LINESTRING", warn = FALSE)
  
  # Remove zero-length geometries
  lengths <- st_length(diff_geom)
  diff_geom <- diff_geom[as.numeric(lengths) > 0.1]  # > 10cm
  
  st_sf(geometry = diff_geom, crs = crs)
}

added_partial <- process_difference(added_raw, crs_work)
removed_partial <- process_difference(removed_raw, crs_work)

# 3.5 Length filtering function (single definition)
len_ok <- function(x, min_len = 10){
  if (!inherits(x, "sf") || nrow(x) == 0) return(x)
  len_m <- as.numeric(st_length(x))
  x <- x[len_m >= min_len, ]
  x
}

added <- len_ok(added_partial, min_len = min_len)
removed <- len_ok(removed_partial, min_len = min_len)

# ==== 4) Debugging: Check what's happening ====
# Check if specific 2023 segments are missing
if (nrow(added) == 0 & nrow(cyc23_n) > 0) {
  cat("Warning: No segments detected as added. Possible issues:\n")
  cat("1. Tolerance too large (current:", tol_m, "m)\n")
  cat("2. CRS mismatch\n")
  cat("3. Geometry issues\n")
  
  # Test: check if buffers are covering everything
  coverage_test <- st_covered_by(st_geometry(cyc23_n), buf19)
  cat("2023 segments fully covered by 2019 buffer:", sum(coverage_test), "/", nrow(cyc23_n), "\n")
}

# ==== 5) Midpoints for quick validation sampling ====
midpoint_one <- function(x) {
  # one sample per line: midpoint
  pts <- st_line_sample(x, sample = 0.5)
  st_as_sf(data.frame(id = seq_along(pts)), geometry = st_cast(pts, "POINT"), crs = crs_work)
}

added_pts   <- midpoint_one(added)   |> mutate(class = "ADD")
removed_pts <- midpoint_one(removed) |> mutate(class = "REMOVE")

# put this before the map chunk (or set eval: true on your boundary chunk)
if (!exists("city_perimeter")) {
  city_perimeter <- sf::st_read("data/barcelona_perimeter.gpkg",
                                layer = "barcelona_perimeter", quiet = TRUE)
}
city_perimeter <- sf::st_transform(city_perimeter, 4326)


library(RColorBrewer)

# Dark2 (no purple); same thickness for all lines
pal <- brewer.pal(8, "Dark2")
col_2019    <- pal[8]         # grey
col_added   <- pal[2]         # orange
col_2023    <- pal[5]         # green (optional toggle)
col_removed <- "#FFFFFF"
col_halo    <- "#000000"

alpha_2019  <- 0.90
alpha_added <- 0.95
alpha_2023  <- 0.75
alpha_halo  <- 0.18

leaflet(options = leafletOptions(preferCanvas = TRUE)) %>%
  addProviderTiles("CartoDB.Positron", group = "Positron") %>%
  addProviderTiles("OpenStreetMap",     group = "OSM") %>%

  # <<< add the boundary before other overlays
  addPolygons(data = city_perimeter, group = "Boundary",
              weight = 2, color = "#222222", fill = FALSE, fillOpacity = 0) %>%

  # your existing layers...
  addPolylines(data = st_transform(cyc19_n, 4326), group = "2019",
               weight = 2, color = col_2019, opacity = alpha_2019) %>%
  addPolylines(data = st_transform(removed, 4326), group = "Removed",
               weight = 2.6, color = col_halo, opacity = alpha_halo) %>%
  addPolylines(data = st_transform(removed, 4326), group = "Removed",
               weight = 2.0, color = col_removed, opacity = 1.0) %>%
  addPolylines(data = st_transform(added, 4326), group = "Added",
               weight = 2, color = col_added, opacity = alpha_added) %>%
  addPolylines(data = st_transform(cyc23_n, 4326), group = "2023",
               weight = 2, color = col_2023, opacity = alpha_2023) %>%

  addLayersControl(
    baseGroups    = c("Positron","OSM"),
    overlayGroups = c("Boundary","2019","Added","Removed","2023"),
    options = layersControlOptions(collapsed = TRUE)
  ) %>%
  hideGroup("2023") %>%
  addLegend(
    position = "bottomright",
    title = "Layers",
    colors = c(col_2019, col_added, col_removed),
    labels = c("2019 (baseline)", "Added (19→23)", "Removed (19→23)"),
    opacity = 1
  )
```

### Sampling frame and design

- Stratified tracts. Pre-selected census tracts by centrality × density (3×3).

- Per-tract quotas (example). ADD=2, REMOVE=1, GENERAL=2 (GENERAL = 2023 non-CI network). Length-weighted sampling within tract; one midpoint per sampled segment.

- Expected N. ~ #tracts × (2+1+2), minus tracts with no candidates in a class.

```{r}
#| include: false
# ================================================================
# STRATIFIED SAMPLING AND STREET VIEW LINK GENERATION
# ================================================================

# ---- Load required libraries ----
library(sf)
library(dplyr)
library(tidyr)
library(biscale)
library(ggplot2)
library(cowplot)
library(osmdata)
library(purrr)
library(jsonlite)
library(httr)
library(tmap)
library(stringr)
library(sf)
library(leaflet)
library(htmltools)
```

```{r}
#| include: false
# ================================================================
# 1. LOAD AND PREPARE DATA
# ================================================================

# Load spatial and population data
tracts_geo <- readRDS("data/bcn-seccio-censal/BCN_seccion_censal_2015.rds")
tracts_pop <- readRDS("data/pop/Population_Census_tract_2015_2022.rds")

tracts_pop_wide <- tracts_pop |> 
  filter(Year %in% c("2015", "2019", "2022")) |> 
  pivot_wider(names_from = Year, values_from = Population, names_prefix = "pop_")

# Merge geometries with population data
barcelona_tracts <- left_join(tracts_geo, tracts_pop_wide, by = c("CUSEC" = "GEOID")) |> 
  st_transform(25831)

# Compute area, population density, and distance to city center
placa_catalunya <- st_sfc(st_point(c(431870, 4581450)), crs = 25831)

barcelona_tracts <- barcelona_tracts |> 
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    dens_2015 = pop_2015 / area_km2,
    dens_2019 = pop_2019 / area_km2,
    dens_2022 = pop_2022 / area_km2,
    centroid = st_centroid(geometry),
    dist_centre_km = as.numeric(st_distance(centroid, placa_catalunya)) / 1000
  )
```

```{r}
#| message: false
#| warning: false
#| include: false
#| paged-print: false

# ================================================================
# 2. STRATIFIED SAMPLING
# ================================================================

barcelona_tracts <- barcelona_tracts |>
  mutate(
    dens_stratum = ntile(dens_2022, 3),
    cent_stratum = ntile(-dist_centre_km, 3),
    stratum_id = paste0("D", dens_stratum, "_C", cent_stratum)
  )

set.seed(123)
per_stratum <- 5  # try 3–6 depending on effort

sampled_tracts <- barcelona_tracts |> 
  group_by(stratum_id) |> 
  sample_n(size = min(per_stratum, n())) |> 
  ungroup()
```

```{r}
#| echo: false

# ================================================================
# 3. BIVARIATE MAPPING
# ================================================================

barcelona_tracts <- barcelona_tracts |> 
mutate(centrality_flipped = -dist_centre_km)
 
bb_bivar <- bi_class(barcelona_tracts, x = dens_2022, y = centrality_flipped, style = "quantile", dim = 3)
 
bivar_map <- ggplot() +
   geom_sf(data = bb_bivar, aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
   bi_scale_fill(pal = "DkBlue2", dim = 3) +
   geom_sf(data = sampled_tracts, fill = NA, color = "black", linewidth = 0.5) +
   labs(title = "") +
   bi_theme() +
   theme(plot.title = element_text(size = 12, hjust = 0.5))
 
 bivar_legend <- bi_legend(pal = "DkBlue2", dim = 3, xlab = "Higher Density", ylab = "More Central", size = 9) +
   theme(axis.title = element_text(size = 9), axis.text = element_blank(),
         plot.margin = margin(5, 5, 5, 5), panel.background = element_blank(), plot.background = element_blank())
 
 final_plot <- ggdraw() +
   draw_plot(bivar_map, 0, 0, 1, 1) +
   draw_plot(bivar_legend, x = 0.73, y = 0.03, width = 0.26, height = 0.26)
 
 final_plot
 
 ggsave("figs/stratified_sample_bivariate_map.png", final_plot, width = 8, height = 8, dpi = 300)
```

### Validation source and coding

- Imagery. Google Street View (GSV).

- Tolerance windows. 2019 target: ±1 year; 2023 target: ±2 years (fixed across coders/cities).

- Per-point fields (Excel).

  + Metadata: class (ADD/REMOVE/GENERAL), tract_id, stratum, lon, lat, gsv_link (+ clickable “Open GSV”).

  + Verifiability: verifiable_2019, verifiable_2023 (TRUE/FALSE).

  + Presence: present_2019, present_2023 (TRUE/FALSE/NA).

  + notes (obstruction, roadworks, poor angle, etc.).

- Consistency rules. If verifiable_* = FALSE → set corresponding present_* = NA. If imagery exists but unclear → keep verifiable_* = TRUE, set present_* = NA.

```{r}
#| eval: false
#| include: false
# ======================= CLEAN VALIDATION PIPELINE ============================
library(sf)
library(dplyr)

# --- 0) Helpers ---------------------------------------------------------------

# Base roads WITHOUT cycle infra (strict complement of your cycle selector)
pick_noncycle_strict <- function(x){
  stopifnot(inherits(x,"sf"))
  nm <- names(x); n <- nrow(x)
  get_chr <- function(col) if (col %in% nm) as.character(x[[col]]) else rep(NA_character_, n)
  highway      <- get_chr("highway")
  bicycle      <- get_chr("bicycle")
  bicycle_road <- get_chr("bicycle_road")
  cycols <- nm[grepl("^cycleway($|:)", nm)]
  has_cycle_tag <- if (length(cycols) > 0) {
    Reduce(`|`, lapply(cycols, function(cn){
      v <- get_chr(cn); !is.na(v) & v != "" & v != "no"
    }))
  } else rep(FALSE, n)
  base_roads <- highway %in% c("trunk","primary","secondary","tertiary",
                               "unclassified","residential","service",
                               "primary_link","secondary_link","tertiary_link")
  keep_cycle <- highway == "cycleway" |
    (highway %in% c("path","track") & bicycle %in% c("designated","yes","permissive")) |
    has_cycle_tag | bicycle_road == "yes" |
    (highway %in% c("living_street","pedestrian") & bicycle == "designated")
  x[base_roads & !keep_cycle, , drop = FALSE]
}

# Midpoint per line (robust to LINESTRING/MULTILINESTRING/collections)
midpoint_one <- function(x){
  if (!inherits(x,"sf") || nrow(x) == 0)
    return(sf::st_sf(id = integer(0), geometry = sf::st_sfc(crs = sf::st_crs(x))))
  g <- sf::st_geometry(x) |>
       sf::st_make_valid() |>
       sf::st_cast("MULTILINESTRING", warn = FALSE) |>
       sf::st_cast("LINESTRING",      warn = FALSE)
  if (length(g) == 0)
    return(sf::st_sf(id = integer(0), geometry = sf::st_sfc(crs = sf::st_crs(x))))
  pts <- sf::st_line_sample(g, sample = 0.5)
  sf::st_as_sf(
    data.frame(id = seq_along(pts)),
    geometry = sf::st_cast(pts, "POINT"),
    crs = sf::st_crs(x)
  )
}

# Length-weighted row indices (safe on zero/NA lengths)
w_sample_idx <- function(x, n){
  if (!inherits(x,"sf") || nrow(x) == 0 || n <= 0) return(integer(0))
  L <- as.numeric(sf::st_length(x)); S <- sum(L, na.rm = TRUE)
  if (!is.finite(S) || S == 0) {
    sample(seq_len(nrow(x)), size = min(n, nrow(x)), replace = FALSE)
  } else {
    sample(seq_len(nrow(x)), size = min(n, nrow(x)), replace = FALSE, prob = L/S)
  }
}

# Points → lon/lat + GSV link (0-row safe; preserves existing columns)
to_lonlat_tbl <- function(pts_sf){
  stopifnot(inherits(pts_sf, "sf"))
  n <- nrow(pts_sf)
  if (n == 0){
    base <- sf::st_drop_geometry(pts_sf)[0, , drop = FALSE]
    base$lon <- numeric(0); base$lat <- numeric(0); base$gsv_link <- character(0)
    return(base)
  }
  wgs    <- sf::st_transform(pts_sf, 4326)
  coords <- sf::st_coordinates(wgs)
  out    <- sf::st_drop_geometry(wgs)
  out$lon <- coords[,1]; out$lat <- coords[,2]
  out$gsv_link <- paste0("https://www.google.com/maps/@?api=1&map_action=pano&viewpoint=",
                         out$lat, ",", out$lon)
  out
}

# Add validation placeholders (NA logical flags + notes) — 0-row safe
add_validation_cols <- function(df){
  n <- nrow(df)
  df$present_2019    <- rep(NA, n)
  df$present_2023    <- rep(NA, n)
  df$verifiable_2019 <- rep(NA, n)
  df$verifiable_2023 <- rep(NA, n)
  df$notes           <- rep(NA_character_, n)
  df
}

# Coerce critical columns so bind_rows never complains
coerce_for_bind <- function(df){
  if (!"class" %in% names(df))    df$class <- character(nrow(df))
  if (!"interval" %in% names(df)) df$interval <- character(nrow(df))
  df$class            <- as.character(df$class)
  df$interval         <- as.character(df$interval)
  if ("gsv_link" %in% names(df)) df$gsv_link <- as.character(df$gsv_link)
  if ("lon" %in% names(df))      df$lon      <- as.numeric(df$lon)
  if ("lat" %in% names(df))      df$lat      <- as.numeric(df$lat)
  if ("present_2019" %in% names(df))    df$present_2019    <- as.logical(df$present_2019)
  if ("present_2023" %in% names(df))    df$present_2023    <- as.logical(df$present_2023)
  if ("verifiable_2019" %in% names(df)) df$verifiable_2019 <- as.logical(df$verifiable_2019)
  if ("verifiable_2023" %in% names(df)) df$verifiable_2023 <- as.logical(df$verifiable_2023)
  if ("notes" %in% names(df))           df$notes           <- as.character(df$notes)
  df
}

# --- 1) Sample sizes ----------------------------------------------------------
set.seed(42)
N_ADD <- min(120, nrow(added))
N_REM <- min(60,  nrow(removed))
N_GEN <- 120

# --- 2) ADDED / REMOVED → midpoints ------------------------------------------
add_idx <- w_sample_idx(added,   N_ADD)
rem_idx <- w_sample_idx(removed, N_REM)

added_pts   <- midpoint_one(added[add_idx, ])   |> dplyr::mutate(class = "ADD")
removed_pts <- midpoint_one(removed[rem_idx, ]) |> dplyr::mutate(class = "REMOVE")

# --- 3) GENERAL (2023 non-cycle) → midpoints ---------------------------------
noncycle23 <- l23 |>
  sf::st_transform(crs_work) |>
  pick_noncycle_strict() |>
  sf::st_make_valid() |>
  sf::st_cast("MULTILINESTRING", warn = FALSE) |>
  sf::st_cast("LINESTRING",      warn = FALSE)

# Drop empties without dplyr pronouns
noncycle23 <- noncycle23[!sf::st_is_empty(sf::st_geometry(noncycle23)), , drop = FALSE]

gen_idx <- w_sample_idx(noncycle23, N_GEN)
gen_pts <- midpoint_one(noncycle23[gen_idx, ]) |> dplyr::mutate(class = "GENERAL")

# --- 4) Tables with placeholders, bind ----------------------------------------
added_tbl   <- to_lonlat_tbl(added_pts)   |> dplyr::mutate(interval = "2019→2023") |> add_validation_cols() |> coerce_for_bind()
removed_tbl <- to_lonlat_tbl(removed_pts) |> dplyr::mutate(interval = "2019→2023") |> add_validation_cols() |> coerce_for_bind()
general_tbl <- to_lonlat_tbl(gen_pts)     |> dplyr::mutate(interval = "2019→2023") |> add_validation_cols() |> coerce_for_bind()

samples_tbl <- dplyr::bind_rows(added_tbl, removed_tbl, general_tbl)

# --- (optional) sanity print --------------------------------------------------
cat("rows → ADD:", nrow(added_tbl), " REMOVE:", nrow(removed_tbl), " GENERAL:", nrow(general_tbl), "\n")
# readr::write_csv(samples_tbl, "samples_2019_2023.csv")
# ===================== END CLEAN VALIDATION PIPELINE ==========================


# 5) Write Excel (3 sheets)
# install.packages("writexl")  # run once if needed
suppressPackageStartupMessages(requireNamespace("writexl", quietly = TRUE))

# ---- Step 5: Excel export with clickable GSV links --------------------------
# install.packages("openxlsx")  # run once if you don't have it
library(openxlsx)
library(dplyr)

outfile <- "samples_2019_2023.xlsx"  # change path/name if you like

# Make a copy with a "GSV" column where we'll write HYPERLINK() formulas
df_xlsx <- samples_tbl %>%
  mutate(GSV = NA_character_) %>%
  relocate(GSV, .after = gsv_link)  # optional positioning

wb <- createWorkbook()
addWorksheet(wb, "validation_samples")
writeData(wb, "validation_samples", df_xlsx)

# Fill the GSV column with clickable hyperlinks (one per row)
n <- nrow(samples_tbl)
if (n > 0) {
  link_col <- which(names(df_xlsx) == "GSV")
  formulas <- paste0('HYPERLINK("', samples_tbl$gsv_link, '","Open GSV")')
  writeFormula(wb, sheet = "validation_samples",
               x = formulas, startCol = link_col, startRow = 2)
}

# Nice-to-haves
addFilter(wb, "validation_samples", row = 1, cols = 1:ncol(df_xlsx))
setColWidths(wb, "validation_samples", cols = 1:ncol(df_xlsx), widths = "auto")
freezePane(wb, "validation_samples", firstActiveRow = 2, firstActiveCol = 1)

saveWorkbook(wb, outfile, overwrite = TRUE)
message("Wrote: ", normalizePath(outfile))


# (Optional) Also write a GPKG with the sampled points for GIS inspection
out_gpkg <- file.path("data","validation_2019_2023_points.gpkg")
if (nrow(added_pts))   sf::st_write(added_pts,   out_gpkg, "Added_pts",   delete_layer = TRUE, quiet = TRUE)
if (nrow(removed_pts)) sf::st_write(removed_pts, out_gpkg, "Removed_pts", delete_layer = TRUE, quiet = TRUE)
if (nrow(gen_pts))     sf::st_write(gen_pts,     out_gpkg, "General_pts", delete_layer = TRUE, quiet = TRUE)
cat("Wrote:", out_gpkg, "\n")
```

```{r}
#| label: helpers-validation
#| echo: false
#| message: false
#| warning: false
#| eval: true
#| include: false

library(sf); library(dplyr)

# 1) One-point-per-row midpoint (no row explosion)
if (!exists("midpoint_one")) {
  midpoint_one <- function(x){
    if (!inherits(x,"sf") || !nrow(x))
      return(sf::st_sf(id=integer(0), geometry=sf::st_sfc(crs=sf::st_crs(x))))
    g <- sf::st_geometry(x) |> sf::st_make_valid() |> suppressWarnings(sf::st_line_merge())
    sf::st_as_sf(data.frame(id=seq_len(nrow(x))),
                 geometry = sf::st_point_on_surface(g),
                 crs = sf::st_crs(x))
  }
}

# 2) Points -> lon/lat + GSV
if (!exists("to_lonlat_tbl")) {
  to_lonlat_tbl <- function(pts_sf){
    stopifnot(inherits(pts_sf,"sf"))
    n <- nrow(pts_sf)
    if (!n) {
      base <- sf::st_drop_geometry(pts_sf)[0,,drop=FALSE]
      base$lon <- numeric(0); base$lat <- numeric(0); base$gsv_link <- character(0)
      return(base)
    }
    wgs <- sf::st_transform(pts_sf, 4326); xy <- sf::st_coordinates(wgs)
    out <- sf::st_drop_geometry(wgs)
    out$lon <- xy[,1]; out$lat <- xy[,2]
    out$gsv_link <- paste0("https://www.google.com/maps/@?api=1&map_action=pano&viewpoint=", out$lat, ",", out$lon)
    out
  }
}

# 3) Validation columns
if (!exists("add_validation_cols")) {
  add_validation_cols <- function(df){
    n <- nrow(df)
    df$present_2019    <- rep(NA, n)
    df$present_2023    <- rep(NA, n)
    df$verifiable_2019 <- rep(NA, n)
    df$verifiable_2023 <- rep(NA, n)
    df$notes           <- rep(NA_character_, n)
    df
  }
}

# 4) Coerce for bind_rows
if (!exists("coerce_for_bind")) {
  coerce_for_bind <- function(df){
    if (!"class" %in% names(df))    df$class <- character(nrow(df))
    if (!"interval" %in% names(df)) df$interval <- character(nrow(df))
    df$class    <- as.character(df$class)
    df$interval <- as.character(df$interval)
    if ("gsv_link" %in% names(df)) df$gsv_link <- as.character(df$gsv_link)
    if ("lon" %in% names(df))      df$lon      <- as.numeric(df$lon)
    if ("lat" %in% names(df))      df$lat      <- as.numeric(df$lat)
    if ("present_2019" %in% names(df))    df$present_2019    <- as.logical(df$present_2019)
    if ("present_2023" %in% names(df))    df$present_2023    <- as.logical(df$present_2023)
    if ("verifiable_2019" %in% names(df)) df$verifiable_2019 <- as.logical(df$verifiable_2019)
    if ("verifiable_2023" %in% names(df)) df$verifiable_2023 <- as.logical(df$verifiable_2023)
    if ("notes" %in% names(df))           df$notes           <- as.character(df$notes)
    df
  }
}

# 5) Non-cycle strict (only real infra counts as cycle)
if (!exists("pick_noncycle_strict")) {
  pick_noncycle_strict <- function(x){
    stopifnot(inherits(x,"sf")); if (!nrow(x)) return(x)
    nm <- names(x); n <- nrow(x)
    gchr <- function(col) if (col %in% nm) tolower(as.character(x[[col]])) else rep(NA_character_, n)
    highway      <- gchr("highway")
    bicycle      <- gchr("bicycle")
    bicycle_road <- gchr("bicycle_road")
    lane_cols <- c("cycleway","cycleway:both","cycleway:left","cycleway:right")
    lane_vals <- c("lane","track","opposite_lane","opposite_track","separate")
    has_lane <- rep(FALSE, n)
    for (col in lane_cols) if (col %in% nm) { v <- gchr(col); has_lane <- has_lane | (v %in% lane_vals) }
    is_cycleway <- highway == "cycleway"
    is_path     <- (highway %in% c("path","footway","track")) & (bicycle %in% c("designated","yes"))
    is_bike_rd  <- bicycle_road == "yes"
    keep_cycle  <- is_cycleway | has_lane | is_path | is_bike_rd
    base_ok <- highway %in% c("trunk","primary","secondary","tertiary",
                              "unclassified","residential","service",
                              "primary_link","secondary_link","tertiary_link",
                              "living_street","pedestrian")
    x[base_ok & !keep_cycle, , drop = FALSE]
  }
}
```


```{r}
#| label: validation-stratified
#| echo: false
#| message: false
#| warning: false
#| eval: true
#| include: false

# safety: define if missing
if (!exists("pick_noncycle_strict")) {
  pick_noncycle_strict <- function(x){
    stopifnot(inherits(x,"sf"))
    if (!nrow(x)) return(x)
    nm <- names(x); n <- nrow(x)
    gchr <- function(col) if (col %in% nm) tolower(as.character(x[[col]])) else rep(NA_character_, n)
    highway      <- gchr("highway")
    bicycle      <- gchr("bicycle")
    bicycle_road <- gchr("bicycle_road")
    lane_cols <- c("cycleway","cycleway:both","cycleway:left","cycleway:right")
    lane_vals <- c("lane","track","opposite_lane","opposite_track","separate")
    has_physical_lane <- rep(FALSE, n)
    for (col in lane_cols) if (col %in% nm) {
      v <- gchr(col); has_physical_lane <- has_physical_lane | (v %in% lane_vals)
    }
    is_cycleway        <- highway == "cycleway"
    is_designated_path <- (highway %in% c("path","footway","track")) & (bicycle %in% c("designated","yes"))
    is_bicycle_road    <- bicycle_road == "yes"
    keep_cycle <- is_cycleway | has_physical_lane | is_designated_path | is_bicycle_road
    base_roads <- highway %in% c("trunk","primary","secondary","tertiary",
                                 "unclassified","residential","service",
                                 "primary_link","secondary_link","tertiary_link",
                                 "living_street","pedestrian")
    x[base_roads & !keep_cycle, , drop = FALSE]
  }
}


# Requires: added, removed, l23, crs_work, sampled_tracts (sf POLYGONS)
stopifnot(exists("added"), exists("removed"), exists("l23"), exists("crs_work"))
stopifnot(exists("sampled_tracts"))

# --- Ensure sampled_tracts has a 'tract_id' ----------------------------------
normalize_tract_id <- function(s){
  nm  <- names(s); nml <- tolower(nm)
  candidates <- c("tract_id","tractid","geoid","geoid10","geoid20","id","code","name","objectid","fid")
  hit <- which(nml %in% candidates)
  if (length(hit) > 0) {
    s$tract_id <- as.character(s[[ nm[hit[1]] ]])
  } else {
    s$tract_id <- sprintf("tract_%04d", seq_len(nrow(s)))
  }
  s
}
sampled_tracts <- normalize_tract_id(sampled_tracts)
stopifnot("tract_id" %in% names(sampled_tracts))

# --- Quotas per tract (tune as needed) ---------------------------------------
PERTRACT_ADD <- 2
PERTRACT_REM <- 1
PERTRACT_GEN <- 2
FORCE_QUOTA  <- FALSE   # TRUE = allow replacement to hit quota

# --- Ensure tracts are in work CRS and have 'stratum' ------------------------
tracts <- sf::st_transform(sampled_tracts, crs_work)
if (!"stratum" %in% names(tracts)) {
  if (all(c("density_f","centrality_f") %in% names(tracts))) {
    tracts$stratum <- interaction(tracts$density_f, tracts$centrality_f, sep = "×", drop = TRUE)
  } else {
    tracts$stratum <- factor("all")
  }
}
tracts$stratum <- droplevels(as.factor(tracts$stratum))

# --- Helper: sample lines within each tract (length-weighted) ----------------
# --- Helper: sample lines within each tract (robust probs) --------------------
sample_lines_by_tract <- function(lines_sf, tracts_sf, n_per, replace = FALSE, min_len = 0){
  if (!inherits(lines_sf,"sf") || nrow(lines_sf) == 0) {
    out <- lines_sf[0, , drop = FALSE]
    out$tract_id <- character(0); out$stratum <- factor()[0]
    return(out)
  }
  L <- sf::st_transform(lines_sf, sf::st_crs(tracts_sf))
  if (min_len > 0) {
    L <- L[as.numeric(sf::st_length(L)) >= min_len, , drop = FALSE]
  }
  if (!nrow(L)) {
    out <- L; out$tract_id <- character(0); out$stratum <- factor()[0]; return(out)
  }

  hits  <- sf::st_intersects(tracts_sf, L)  # per-tract list of row indices
  picks <- vector("list", length = nrow(tracts_sf))

  for (i in seq_len(nrow(tracts_sf))) {
    idx <- hits[[i]]
    if (!length(idx)) next

    # how many to draw for this tract
    draw_with_replacement <- isTRUE(replace) && length(idx) < n_per
    k <- if (draw_with_replacement) n_per else min(n_per, length(idx))

    # local indices (1..length(idx)) and length weights
    local <- seq_along(idx)
    lens  <- as.numeric(sf::st_length(L[idx, , drop = FALSE]))
    lens[!is.finite(lens)] <- 0
    sprob <- sum(lens, na.rm = TRUE)

    # only pass prob if valid and matching local length
    prob <- if (sprob > 0 && length(lens) == length(local)) lens / sprob else NULL

    pick_local <- sample(local, size = k, replace = draw_with_replacement, prob = prob)
    pick_idx   <- idx[pick_local]

    piece <- L[pick_idx, , drop = FALSE]
    piece$tract_id <- tracts_sf$tract_id[i]
    piece$stratum  <- tracts_sf$stratum[i]
    picks[[i]] <- piece
  }
  dplyr::bind_rows(picks)
}

# --- GENERAL (2023 non-cycle) for FN test ------------------------------------
# --- GENERAL pool (2023 non-cycle): tag-based, then spatial fallback if empty ---
noncycle23 <- l23 |>
  sf::st_transform(crs_work) |>
  pick_noncycle_strict() |>
  sf::st_make_valid() |>
  sf::st_cast("MULTILINESTRING", warn = FALSE) |>
  sf::st_cast("LINESTRING",      warn = FALSE)
if (nrow(noncycle23)) {
  noncycle23 <- noncycle23[!sf::st_is_empty(sf::st_geometry(noncycle23)), , drop = FALSE]
}

# Fallback if tag-based pool is empty
if (!nrow(noncycle23)) {
  base_roads <- c("trunk","primary","secondary","tertiary",
                  "unclassified","residential","service",
                  "primary_link","secondary_link","tertiary_link",
                  "living_street","pedestrian")

  # 1) Base roads (lines only)
  base23 <- l23 |>
    sf::st_transform(crs_work) |>
    dplyr::filter(.data$highway %in% base_roads)
  keep_lines <- sf::st_geometry_type(base23) %in% c("LINESTRING","MULTILINESTRING")
  base23 <- base23[keep_lines, , drop = FALSE] |>
            sf::st_make_valid() |>
            suppressWarnings(sf::st_cast("LINESTRING"))

  # 2) Buffer 2023 cycle network
  cyc_geom <- NULL
  if (exists("cyc23_n") && inherits(cyc23_n,"sf") && nrow(cyc23_n)) {
    cyc_geom <- sf::st_buffer(sf::st_geometry(cyc23_n), tol_m) |> sf::st_union()
  } else if (exists("cyc23") && inherits(cyc23,"sf") && nrow(cyc23)) {
    cyc_geom <- sf::st_buffer(sf::st_geometry(sf::st_transform(cyc23, crs_work)), tol_m) |> sf::st_union()
  }

  # 3) Difference and keep only lines
  if (!is.null(cyc_geom) && length(cyc_geom) > 0) {
    diff_geom <- sf::st_difference(sf::st_geometry(base23), cyc_geom)
    diff_geom <- sf::st_collection_extract(diff_geom, "LINESTRING", warn = FALSE)
    noncycle23 <- sf::st_sf(geometry = diff_geom, crs = crs_work)
    if (nrow(noncycle23)) {
      noncycle23 <- noncycle23[!sf::st_is_empty(sf::st_geometry(noncycle23)), , drop = FALSE]
    }
  } else {
    noncycle23 <- base23
  }
}
# --- STRATIFIED PICKS (per-tract quotas) ---
added_by_tr   <- sample_lines_by_tract(added,      tracts, PERTRACT_ADD, replace = FORCE_QUOTA)
removed_by_tr <- sample_lines_by_tract(removed,    tracts, PERTRACT_REM, replace = FORCE_QUOTA)
# Keep GENERAL only for tracts where we already sampled ADD or REMOVE
# After you build added_by_tr / removed_by_tr:
sampled_ids <- unique(added_by_tr$tract_id)  # <-- ADD-only

tracts_for_general <- dplyr::filter(tracts, tract_id %in% sampled_ids)
if (nrow(tracts_for_general) == 0) tracts_for_general <- tracts  # fallback

general_by_tr <- sample_lines_by_tract(
  noncycle23, tracts_for_general, PERTRACT_GEN, replace = FORCE_QUOTA
)

# --- Midpoints (1 point per sampled row) + metadata ---
# If you already defined midpoint_one elsewhere, keep that; otherwise use this:
midpoint_one <- function(x){
  if (!inherits(x,"sf") || !nrow(x)) return(sf::st_sf(id=integer(0), geometry=sf::st_sfc(crs=sf::st_crs(x))))
  g <- sf::st_geometry(x) |> sf::st_make_valid() |> suppressWarnings(sf::st_line_merge())
  sf::st_as_sf(data.frame(id=seq_len(nrow(x))),
               geometry = sf::st_point_on_surface(g),
               crs = sf::st_crs(x))
}

added_pts <- midpoint_one(added_by_tr)   |> dplyr::mutate(class="ADD",    tract_id=added_by_tr$tract_id,   stratum=added_by_tr$stratum)
removed_pts <- midpoint_one(removed_by_tr)|> dplyr::mutate(class="REMOVE", tract_id=removed_by_tr$tract_id, stratum=removed_by_tr$stratum)
gen_pts   <- midpoint_one(general_by_tr) |> dplyr::mutate(class="GENERAL", tract_id=general_by_tr$tract_id, stratum=general_by_tr$stratum)

# --- Tables with lon/lat + placeholders, then bind ---------------------------
add_tbl <- to_lonlat_tbl(added_pts)   |> dplyr::mutate(interval = "2019→2023") |> add_validation_cols() |> coerce_for_bind()
rem_tbl <- to_lonlat_tbl(removed_pts) |> dplyr::mutate(interval = "2019→2023") |> add_validation_cols() |> coerce_for_bind()
gen_tbl <- to_lonlat_tbl(gen_pts)     |> dplyr::mutate(interval = "2019→2023") |> add_validation_cols() |> coerce_for_bind()

samples_tbl <- dplyr::bind_rows(add_tbl, rem_tbl, gen_tbl)
```

```{r}
#| label: tables-export
#| echo: false
#| message: false
#| warning: false
#| eval: true

stopifnot(exists("add_tbl"), exists("rem_tbl"), exists("gen_tbl"))

# Where to save
outdir  <- "outputs"
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
outfile <- file.path(outdir, paste0(city_tag, "_samples_2019_2023.xlsx"))

library(openxlsx)
library(dplyr)

wb <- createWorkbook()

write_one <- function(df, sheet){
  df2 <- df %>%
    mutate(GSV = NA_character_) %>%
    relocate(GSV, .after = gsv_link)
  addWorksheet(wb, sheet)
  writeData(wb, sheet, df2)
  if (nrow(df2) > 0) {
    link_col <- which(names(df2) == "GSV")
    writeFormula(
      wb, sheet,
      x = paste0('HYPERLINK("', df$gsv_link, '","Open GSV")'),
      startCol = link_col, startRow = 2
    )
  }
  addFilter(wb, sheet, row = 1, cols = 1:ncol(df2))
  setColWidths(wb, sheet, cols = 1:ncol(df2), widths = "auto")
  freezePane(wb, sheet, firstActiveRow = 2, firstActiveCol = 1)
}

write_one(add_tbl, "ADD")
write_one(rem_tbl, "REMOVE")
write_one(gen_tbl, "GENERAL")

saveWorkbook(wb, outfile, overwrite = TRUE)
message("Excel written: ", normalizePath(outfile))
```

```{r}
#| label: validation-map-points
#| echo: false
#| message: false
#| warning: false

library(leaflet)
library(sf)
library(dplyr)

# ---- helper: prep points (WGS84 + lon/lat + popup) --------------------------
prep_pts <- function(x){
  if (!inherits(x,"sf") || !nrow(x)) return(NULL)
  w  <- st_transform(x, 4326)
  xy <- st_coordinates(w)
  good <- is.finite(xy[,1]) & is.finite(xy[,2])
  if (!any(good)) return(NULL)
  w   <- w[good, , drop = FALSE]
  xy  <- xy[good, , drop = FALSE]
  w$lon <- xy[,1]; w$lat <- xy[,2]
  w$gsv <- sprintf("https://www.google.com/maps/@?api=1&map_action=pano&viewpoint=%f,%f", xy[,2], xy[,1])
  w$popup <- paste0(
    "<b>", w$class, "</b>",
    "<br/>tract: ", w$tract_id,
    if ("stratum" %in% names(w)) paste0("<br/>stratum: ", w$stratum) else "",
    "<br/><a href='", w$gsv, "' target='_blank'>Open GSV</a>"
  )
  w
}

add_wgs <- prep_pts(added_pts)
rem_wgs <- prep_pts(removed_pts)
gen_wgs <- prep_pts(gen_pts)

# ---- bounds (prefer city_perimeter -> points -> tracts -> fallback) ----------
if (exists("city_perimeter")) {
  bb <- st_bbox(st_transform(city_perimeter, 4326))
  bounds <- unname(c(bb$xmin, bb$ymin, bb$xmax, bb$ymax))
} else {
  bbs <- list()
  if (!is.null(add_wgs) && nrow(add_wgs)) bbs <- c(bbs, list(st_bbox(add_wgs)))
  if (!is.null(rem_wgs) && nrow(rem_wgs)) bbs <- c(bbs, list(st_bbox(rem_wgs)))
  if (!is.null(gen_wgs) && nrow(gen_wgs)) bbs <- c(bbs, list(st_bbox(gen_wgs)))
  if (!length(bbs) && exists("tracts") && nrow(tracts)) bbs <- list(st_bbox(st_transform(tracts, 4326)))
  bounds <- if (length(bbs)) {
    M <- do.call(rbind, lapply(bbs, function(b) unname(c(b$xmin,b$ymin,b$xmax,b$ymax))))
    c(min(M[,1]), min(M[,2]), max(M[,3]), max(M[,4]))
  } else c(2.05, 41.30, 2.25, 41.45)  # BCN-ish fallback
}

# ---- colours -----------------------------------------------------------------
# points (Dark2-ish, no purple)
col_add_pt <- "#E6AB02"  # gold
col_rem_pt <- "#D95F02"  # orange-red
col_gen_pt <- "#1B9E77"  # teal
stroke     <- "#000000"

# tracts
tract_stroke <- "#66666666"
tract_fill   <- "#888888"

# networks (optional)
col_2019_net   <- "#666666"
col_added_net  <- "#D95F02"
col_removedNet <- "#FFFFFF"
col_removedHalo<- "#000000"
col_2023_net   <- "#1B9E77"

# ---- map ---------------------------------------------------------------------
m <- leaflet(options = leafletOptions(preferCanvas = TRUE)) %>%
  addProviderTiles("CartoDB.Positron", group = "Positron")

# Tracts polygons (subtle)
if (exists("tracts") && inherits(tracts,"sf") && nrow(tracts)) {
  m <- m %>%
    addPolygons(data = st_transform(tracts, 4326), group = "Tracts",
                weight = 1, color = tract_stroke,
                fillColor = tract_fill, fillOpacity = 0.05, opacity = 0.7)

  # Area labels = tract_id (small, semi-transparent)
  tract_centers <- st_point_on_surface(st_transform(tracts, 4326))
  m <- m %>%
    addCircleMarkers(
      data = tract_centers, group = "Area labels",
      radius = 1, stroke = FALSE, fillOpacity = 0,
      label = ~as.character(tract_id),
      labelOptions = labelOptions(
        noHide = TRUE, direction = "center",
        style = list(
          "font-weight" = "600",
          "font-size"   = "10px",
          "color"       = "rgba(0,0,0,0.45)",
          "text-shadow" = "0 0 2px rgba(255,255,255,0.6)"
        )
      )
    )
}

# Optional network context (added but hidden by default below)
if (exists("cyc19_n") && inherits(cyc19_n,"sf") && nrow(cyc19_n))
  m <- m %>% addPolylines(data = st_transform(cyc19_n, 4326), group = "2019 net",
                          weight = 2, color = col_2019_net, opacity = 0.7)

if (exists("removed") && inherits(removed,"sf") && nrow(removed)) {
  m <- m %>% addPolylines(data = st_transform(removed, 4326), group = "Removed net",
                          weight = 2.6, color = col_removedHalo, opacity = 0.18)
  m <- m %>% addPolylines(data = st_transform(removed, 4326), group = "Removed net",
                          weight = 2.0, color = col_removedNet,  opacity = 1.0)
}

if (exists("added") && inherits(added,"sf") && nrow(added))
  m <- m %>% addPolylines(data = st_transform(added, 4326), group = "Added net",
                          weight = 2, color = col_added_net, opacity = 0.95)

if (exists("cyc23_n") && inherits(cyc23_n,"sf") && nrow(cyc23_n))
  m <- m %>% addPolylines(data = st_transform(cyc23_n, 4326), group = "2023 net",
                          weight = 2, color = col_2023_net, opacity = 0.6)

# Points (explicit lng/lat; active by default)
if (!is.null(add_wgs) && nrow(add_wgs) > 0)
  m <- m %>% addCircleMarkers(data = add_wgs, lng = ~lon, lat = ~lat, group = "ADD pts",
                              radius = 6, weight = 1.5, color = stroke,
                              fillColor = col_add_pt, fillOpacity = 0.95,
                              popup = ~popup)

if (!is.null(rem_wgs) && nrow(rem_wgs) > 0)
  m <- m %>% addCircleMarkers(data = rem_wgs, lng = ~lon, lat = ~lat, group = "REMOVE pts",
                              radius = 6, weight = 1.5, color = stroke,
                              fillColor = col_rem_pt, fillOpacity = 0.95,
                              popup = ~popup)

if (!is.null(gen_wgs) && nrow(gen_wgs) > 0)
  m <- m %>% addCircleMarkers(data = gen_wgs, lng = ~lon, lat = ~lat, group = "GENERAL pts",
                              radius = 6, weight = 1.5, color = stroke,
                              fillColor = col_gen_pt, fillOpacity = 0.95,
                              popup = ~popup)

# Control + single legend
m %>%
  clearControls() %>%
  addLayersControl(
    baseGroups = c("Positron"),
    overlayGroups = c(
      "Tracts","Area labels",
      "2019 net","Added net","Removed net","2023 net",
      "ADD pts","REMOVE pts","GENERAL pts"
    ),
    options = layersControlOptions(collapsed = TRUE)
  ) %>%
  hideGroup(c("2019 net","Added net","Removed net","2023 net")) %>%  # networks start hidden
  fitBounds(bounds[1], bounds[2], bounds[3], bounds[4]) %>%
  addLegend(
    position = "bottomright", title = "Validation",
    colors = c(tract_fill, col_add_pt, col_rem_pt, col_gen_pt),
    labels = c("Tracts (areas)", "ADD points", "REMOVE points", "GENERAL points"),
    opacity = 1
  )

```


### Outcomes and metrics

- ADD precision:

  + TP_ADD = (class=ADD & verif_2023=TRUE & present_2023=TRUE)

  + FP_ADD = (class=ADD & verif_2023=TRUE & present_2023=FALSE)

  + Precision_ADD = TP_ADD / (TP_ADD + FP_ADD)

- REMOVE precision: analogous, with present_2023=FALSE for TP.

- False negatives (GENERAL):

  + Strict FN: (class=GENERAL) & verif_2019=TRUE & present_2019=FALSE & verif_2023=TRUE & present_2023=TRUE.

  + FN_rate = FN_strict / N_strict, where N_strict = GENERAL rows with both years verifiable.

- Recall for additions: Recall_ADD = TP_ADD / (TP_ADD + FN_strict).

- Calibration of OSM km (adds):

$$
K_{\text{add, validated}}
\approx
K_{\text{add, OSM}}
\times
\frac{1 - \mathrm{FP}_{\text{ADD}}}{1 - \mathrm{FN}_{\text{ADD}}}
$$ {#eq-kadd}

(Analogous down-adjustment for removals using FP_REMOVE.)

<!-- Statistical analysis -->

<!-- Report proportions with 95% CIs (e.g., prop.test). -->

<!-- By-stratum estimates (centrality × density); compute a weighted overall (equal weights if equal per stratum; otherwise use design weights reflecting tract/pop coverage). -->

<!-- Sensitivity analyses: -->

<!-- “Loose FN”: verif_2023=TRUE & present_2023=TRUE (regardless of 2019). -->

<!-- Alternative tolerance windows (±0/±1 year). -->

<!-- Inter-rater reliability: double-code ~10% of points; report κ. -->

<!-- Reproducibility and ethics -->

<!-- Provide code (R/Quarto) and exact tag list; no personal data; imagery used is publicly accessible via GSV. -->

<!-- Results (template to fill once coded) -->

<!-- Sample flow. Counts by class and verifiability (table). -->

<!-- ADD: TP/FP counts, Precision_ADD with CI; by-stratum differences (heatmap/table). -->

<!-- REMOVE: TP/FP counts, Precision_REMOVE with CI; by-stratum. -->

<!-- GENERAL: FN_strict and FN_rate with CI; Recall_ADD with CI. -->

<!-- Calibration. Raw OSM km vs validated km (point estimate + CI); % change after adjustment. -->

<!-- Sensitivity. Changes under loose FN and alternative windows. -->

<!-- Reliability. κ for double-coded subset. -->

<!-- Discussion -->

<!-- Principal findings. (e.g., OSM precision high for ADD, moderate for REMOVE; FN_rate non-trivial in peripheral strata; recall X%.) -->

<!-- Implications. When and where OSM is dependable for monitoring CI; how to adjust OSM-based indicators. -->

<!-- Biases. Systematic error by urban context (density/centrality), imagery gaps. -->

<!-- Strengths & limitations. Stratified design; portable workflow; GSV timing constraints; sample size. -->

<!-- Relation to prior work. How estimates compare with municipal reports / prior validations. -->

<!-- Conclusions -->

<!-- OSM is [adequate / limited] for detecting CI changes when [conditions]. -->

<!-- Provide validated growth estimates with uncertainty; recommend routine stratified validation for cross-city comparability. -->

<!-- Tables & Figures (recommended) -->

<!-- T1. Sample counts by class × verifiability (and by stratum). -->

<!-- T2. ADD/REMOVE precision with 95% CIs (overall + by stratum). -->

<!-- T3. GENERAL FN_rate and Recall_ADD with 95% CIs (overall + by stratum). -->

<!-- F1. Stratified sampling map (tracts + sampled points). -->

<!-- F2. Heatmap of metrics by stratum. -->

<!-- F3. Raw vs validated km (adds/removals) with CIs (bar/CI plot). -->

<!-- Supp. Coding protocol examples (what counts as CI), sensitivity plots. -->

<!-- ## References -->
