---
# title: "ATRAPA Work Package 1: Built Environment Transformations Dataset"
format: html
editor: visual
bibliography: wp1.bib
---

🔗 Part of the [ATRAPA database project](https://github.com/GEMOTT/atrapa_database)\

⬅️ [Back to project overview](https://github.com/GEMOTT/atrapa%20database) ➡️ [Next repo related: Electoral and socioeconomic data](https://github.com/GEMOTT/electoral-socioeconomic-data)

# Can OpenStreetMap Reliably Track Changes in Active Travel Infrastructure? Evidence from Barcelona with GSV Validation

## Introduction

The relationship between the built environment and travel behaviour has been widely studied, with many studies identifying associations between environmental characteristics and travel patterns [@cerin_neighbourhood_2017; @ding_neighborhood_2011; @zhang_impact_2022]. However, most research relies on cross-sectional data, which cannot establish causality [@mccormack_search_2011; @coevering_multi-period_2015]. In contrast, studies that track changes in both travel behaviour and the built environment—such as longitudinal studies and natural experiments—offer stronger causal insights but remain relatively scarce [@karmeniemi_built_2018; @smith_systematic_2017; @tcymbal_effects_2020].

One of the main challenges in expanding this area of research is the limited availability of consistent, time-series data on the built environment. While historical data on travel behaviour is often more accessible—through sources like censuses, surveys, and increasingly, crowdsourced platforms like Strava—comparable records of past urban infrastructure are much harder to obtain. Long-term records of active travel networks, though consistent and accessible historical data remains limited and varies across cities, which hinders broader or international comparisons. An alternative is to reconstruct historical built environment data manually using maps, satellite imagery, and planning records, but this process is highly resource-intensive and typically limited in scale.

The growing availability of Volunteered Geographic Information (VGI) presents new opportunities to overcome data limitations in built environment research. Among these sources, OpenStreetMap (OSM) stands out for providing open, editable, and historical data on various types of infrastructure, making it a promising tool for analysing urban transformations over time. However, its application in this context requires careful validation due to well-documented limitations in accuracy, completeness, and temporal consistency [@barron_comprehensive_2014].

While OSM has been widely used for mapping infrastructure and supporting routing applications, its utility for analysing changes in infrastructure over time is less well established. This study seeks to evaluate how accurately historical OSM data reflects changes in active travel infrastructure—specifically bike lanes, pedestrian streets, and living streets. We propose and apply a semi-automated validation method that compares reported OSM changes against external reference sources, including street-level imagery (Google Street View), satellite imagery, and official municipal records.

Focusing on the city of Barcelona, our approach uses stratified sampling to ensure spatial and socio-demographic diversity. While the analysis is limited to one city, the proposed framework is designed to be scalable and transferable, offering a practical methodology for researchers and planners seeking to monitor infrastructure change over time using open data sources.

This study builds on recent efforts to assess OSM’s data quality and potential for infrastructure analysis, with particular attention to its capacity to represent change over time.

## Data and Method

### Data Sources

-   OpenStreetMap (OSM) snapshots: 2015, 2019, 2023

-   Google Street View (GSV) imagery

### Sampling Strategy

-   Unit of analysis: Census tracts (80 in total, Barcelona, 2015 boundaries)

-   Stratification variables:

    -   Population density (2022 estimates)

    -   Centrality: Euclidean distance from Plaça Catalunya, the city center

### Change Detection in OSM

-   Extract infrastructure from OSM for 2015, 2019, 2023

    -   Focus tags: highway=cycleway, cycleway=\*, highway=pedestrian, highway=living_street

-   Compare time periods:

    -   Period 1: 2015–2019

    -   Period 2: 2019–2023

-   Identify changes:

    -   Additions

    -   Removals

    -   Reclassifications

### Validation Strategy

#### OSM-Reported Changes (False Positives)

-   Use GSV imagery and satellite views to verify all detected changes

-   Label each as:

    -   ✅ Confirmed

    -   ❌ False Positive +❓ Uncertain

#### Missed Changes (False Negatives)

-   Sample \~100 street segments from within census tract sample

-   For each segment, check imagery to see if:

    -   Infrastructure exists in reality

    -   But is missing from OSM

### Evaluation Metrics

-   Accuracy = Confirmed OSM changes / All OSM-reported changes

-   Completeness = Confirmed OSM changes / (Confirmed + Missed changes)

-   SCI (Spatial Completeness Index) = Variation in completeness across tracts (e.g., standard deviation)

### Inclusion Criteria

-   Include city/period/type only if:

    -   Completeness ≥ 80%

    -   SCI ≤ 15%

-   Based on prior studies:

    -   Hochmair et al. (2014), Barron et al. (2014), Elwood & Goodchild (2013)

#### Example Evaluation Table by Interval and Infrastructure Type

| City \| Interval \| Type \| Completeness \| SCI \| Accuracy \| Decision \|

\|-----------\|-----------\|-----------\|-----------\|-----------\|-----------\|-----------\|

| Barcelona \| 2015–2019 \| Bike Lanes \| 88% ✅ \| 9% ✅ \| 91% ✅ \| ✅ Include \|

| Barcelona \| 2015–2019 \| Pedestrian \| 84% ✅ \| 12% ✅ \| 87% ✅ \| ✅ Include \|

| Barcelona \| 2015–2019 \| Living Streets \| 72% ❌ \| 18% ❌ \| 78% ❌ \| ❌ Exclude \|

## Results

### Descriptive analyses --\> <!--#

## OSM validation method

### Aim

Validate OSM cycling infrastructure across years (e.g. 2015, 2019, 2023).

### Design

-   Stratified random sample of census tracts (centrality × density, 3×3).

-   Five tracts per stratum (45 in total).

-   Per tract and interval (2015→2019, 2019→2023):

    -   3 random points on the OSM cycleway network (*if present in tract*) → check for false positives for that year.

    -   3 random points on the OSM general road network → check for false negative for that year.

-   Validation source: Google Street View (GSV).

```{r}
#| include: false
# ================================================================
# STRATIFIED SAMPLING AND STREET VIEW LINK GENERATION
# ================================================================

# ---- Load required libraries ----
library(sf)
library(dplyr)
library(tidyr)
library(biscale)
library(ggplot2)
library(cowplot)
library(osmdata)
library(purrr)
library(jsonlite)
library(httr)
library(tmap)
library(stringr)
library(sf)
library(leaflet)
library(htmltools)
```

```{r}
#| include: false
# ================================================================
# 1. LOAD AND PREPARE DATA
# ================================================================

# Load spatial and population data
tracts_geo <- readRDS("data/bcn-seccio-censal/BCN_seccion_censal_2015.rds")
tracts_pop <- readRDS("data/pop/Population_Census_tract_2015_2022.rds")

tracts_pop_wide <- tracts_pop |> 
  filter(Year %in% c("2015", "2019", "2022")) |> 
  pivot_wider(names_from = Year, values_from = Population, names_prefix = "pop_")

# Merge geometries with population data
barcelona_tracts <- left_join(tracts_geo, tracts_pop_wide, by = c("CUSEC" = "GEOID")) |> 
  st_transform(25831)

# Compute area, population density, and distance to city center
placa_catalunya <- st_sfc(st_point(c(431870, 4581450)), crs = 25831)

barcelona_tracts <- barcelona_tracts |> 
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    dens_2015 = pop_2015 / area_km2,
    dens_2019 = pop_2019 / area_km2,
    dens_2022 = pop_2022 / area_km2,
    centroid = st_centroid(geometry),
    dist_centre_km = as.numeric(st_distance(centroid, placa_catalunya)) / 1000
  )
```

```{r}
#| message: false
#| warning: false
#| include: false
#| paged-print: false

# ================================================================
# 2. STRATIFIED SAMPLING
# ================================================================

barcelona_tracts <- barcelona_tracts |>
  mutate(
    dens_stratum = ntile(dens_2022, 3),
    cent_stratum = ntile(-dist_centre_km, 3),
    stratum_id = paste0("D", dens_stratum, "_C", cent_stratum)
  )

set.seed(123)
per_stratum <- 5  # try 3–6 depending on effort

sampled_tracts <- barcelona_tracts |> 
  group_by(stratum_id) |> 
  sample_n(size = min(per_stratum, n())) |> 
  ungroup()
```

### Maps

-   **Map A:** Stratified areas selected.

```{r}
#| echo: false

# ================================================================
# 3. BIVARIATE MAPPING
# ================================================================

barcelona_tracts <- barcelona_tracts |> 
mutate(centrality_flipped = -dist_centre_km)
 
bb_bivar <- bi_class(barcelona_tracts, x = dens_2022, y = centrality_flipped, style = "quantile", dim = 3)
 
bivar_map <- ggplot() +
   geom_sf(data = bb_bivar, aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
   bi_scale_fill(pal = "DkBlue2", dim = 3) +
   geom_sf(data = sampled_tracts, fill = NA, color = "black", linewidth = 0.5) +
   labs(title = "") +
   bi_theme() +
   theme(plot.title = element_text(size = 12, hjust = 0.5))
 
 bivar_legend <- bi_legend(pal = "DkBlue2", dim = 3, xlab = "Higher Density", ylab = "More Central", size = 9) +
   theme(axis.title = element_text(size = 9), axis.text = element_blank(),
         plot.margin = margin(5, 5, 5, 5), panel.background = element_blank(), plot.background = element_blank())
 
 final_plot <- ggdraw() +
   draw_plot(bivar_map, 0, 0, 1, 1) +
   draw_plot(bivar_legend, x = 0.73, y = 0.03, width = 0.26, height = 0.26)
 
 final_plot
 
 ggsave("figs/stratified_sample_bivariate_map.png", final_plot, width = 8, height = 8, dpi = 300)
```

```{r}
#| eval: false
#| include: false
# ================================================================
# 4. NETWORK POINT SAMPLING FROM OSM (MIDPOINT-BASED)
# ================================================================

# Setup
x_fp <- 2  # points per tract from cycleways
x_fn <- 2  # points per tract from general road network
target_crs <- 25831
gsv_api_key <- "[REDACTED_GOOGLE_API_KEY]"
sampled_tracts <- st_transform(sampled_tracts, 25831)

# Limit to 3 tracts and assign tract IDs
sampled_tracts <- st_transform(sampled_tracts, target_crs)
# sampled_tracts <- sampled_tracts[1:10, ]
sampled_tracts$tract_id <- 1:nrow(sampled_tracts)

# Load or read OSM data
osm_file <- "data/osm_streets_barcelona.rds"
osm_streets <- if (!file.exists(osm_file)) {
  opq("Barcelona") %>%
    add_osm_feature(key = "highway") %>%
    osmdata_sf() %>%
    { saveRDS(., osm_file); . }
} else {
  readRDS(osm_file)
}

streets <- st_transform(osm_streets$osm_lines, 25831)

required_cols <- c("cycleway", "cycleway:left", "cycleway:right", "bicycle", "route") # Fill missing OSM columns
streets[setdiff(required_cols, names(streets))] <- NA

cycleway_lines <- streets %>% # Classify lines
  filter(
    highway == "cycleway" |
    cycleway %in% c("lane", "track", "separate", "opposite", "opposite_lane", "opposite_track") |
    `cycleway:left` %in% c("lane", "track", "separate", "opposite", "opposite_lane", "opposite_track") |
    `cycleway:right` %in% c("lane", "track", "separate", "opposite", "opposite_lane", "opposite_track")
  )

general_lines <- streets %>%
  filter(highway %in% c("primary", "secondary", "tertiary", "residential"))

# Test which sampled tracts have general infrastructure
sampled_tracts <- sampled_tracts %>%
  mutate(has_general = lengths(st_intersects(., general_lines)) > 0)

# Optional: inspect how many do/don't
table(sampled_tracts$has_general)

# Test which sampled tracts have general infrastructure
sampled_tracts <- sampled_tracts %>%
  mutate(has_cycleway = lengths(st_intersects(., cycleway_lines)) > 0)

# Optional: inspect how many do/don't
table(sampled_tracts$has_cycleway)

# Sampling function (minimal and working)
sample_points_from_lines <- function(lines, tract_geom, tract_id, n_points, source_label) {
  # Clip lines to the tract (so midpoints are guaranteed inside)
  lines_in <- tryCatch(
    suppressWarnings(st_intersection(lines, tract_geom)),
    error = function(e) lines[0, ]
  )
  # Keep only LINESTRING/MULTILINESTRING parts
  lines_in <- st_collection_extract(lines_in, "LINESTRING")
  if (nrow(lines_in) == 0) return(NULL)

  # Sample many vertices along lines, then pick n_points
  # (use a metric CRS for stable sampling—you're already in EPSG:25831)
  sampled <- tryCatch(
    st_line_sample(st_geometry(lines_in), n = max(n_points * 10, 10), type = "random"),
    error = function(e) st_sfc()
  )
  if (length(sampled) == 0) return(NULL)

  sampled <- st_cast(sampled, "POINT")
  if (length(sampled) > n_points) {
    sampled <- sampled[sample(seq_along(sampled), n_points)]
  }

  st_sf(
    data.frame(tract_id = tract_id, source = source_label),
    geometry = sampled,
    crs = st_crs(tract_geom)
  )
}

# Modified mapping process with better error handling
samples_all <- purrr::map_dfr(1:nrow(sampled_tracts), function(i) {
  tract <- sampled_tracts[i, ]
  geom <- st_geometry(tract)
  
  # ⬇️ OVERSAMPLE HERE — increase from 2 to e.g., 30
  out1 <- sample_points_from_lines(cycleway_lines, geom, tract$tract_id, 10, "osm_cycleway")
  out2 <- sample_points_from_lines(general_lines, geom, tract$tract_id, 10, "osm_general")
  
  # Debugging output
  print(paste("Tract", tract$tract_id, 
              "| Cycleway lines in:", ifelse(is.null(out1), 0, nrow(out1)),
              "| General lines in:", ifelse(is.null(out2), 0, nrow(out2))))
  
  # Safely combine results
  result <- tryCatch({
    if (!is.null(out1) && !is.null(out2)) {
      bind_rows(out1, out2)
    } else if (!is.null(out1)) {
      out1
    } else if (!is.null(out2)) {
      out2
    } else {
      NULL
    }
  }, error = function(e) {
    warning(paste("Error combining results for tract", tract$tract_id, ":", e$message))
    NULL
  })
  
  return(result)
}, .id = "tract_num")

# id_photo not appearing properly
samples_all <- samples_all %>%
  mutate(photo_id = row_number())
```

```{r}
#| include: false
# ================================================================
# 5. ADD GSV DATE + GENERATE LINK
# ================================================================
samples_all_wgs <- st_transform(samples_all, 4326)

get_gsv_date <- function(lat, lon, key) {
  url <- paste0("https://maps.googleapis.com/maps/api/streetview/metadata?",
                "location=", lat, ",", lon, "&key=", key)
  tryCatch({
    res <- jsonlite::fromJSON(url)
    if (!is.null(res$date)) res$date else NA_character_
  }, error = function(e) NA_character_)
}

# ---- Counter-based stop (add this block) ----
counter <- 0L           # resets each run/knit
MAX_CALLS <- 480        # keep some headroom under your 500/day quota

safe_get_gsv_date <- function(lat, lon, key){
  if (counter >= MAX_CALLS) return(NA_character_)
  counter <<- counter + 1L   # use <<- so it updates the outer 'counter'
  get_gsv_date(lat, lon, key)
}
# ---------------------------------------------

samples_all_wgs <- samples_all_wgs %>%
  dplyr::mutate(
    gsv_date = purrr::map_chr(seq_len(n()), function(i) {
      coords <- sf::st_coordinates(samples_all_wgs[i, ])
      safe_get_gsv_date(coords[2], coords[1], gsv_api_key)   # <-- use SAFE version
    }),
    gsv_maps_url = purrr::map_chr(seq_len(n()), function(i) {
      coords <- sf::st_coordinates(samples_all_wgs[i, ])
      paste0("https://www.google.com/maps/@?api=1&map_action=pano&viewpoint=",
             coords[2], ",", coords[1])
    })
  )

samples_all_wgs <- samples_all_wgs %>%
  left_join(
    st_drop_geometry(sampled_tracts)[, c("tract_id", "stratum_id")],
    by = "tract_id"
  )

saveRDS(samples_all_wgs, "data/samples_all_wgs.rds")
```

```{r}
#| include: false
# Select 3 per tract × source (pref ≥2023) (save to RDS)

# If you didn't run the API chunk, create a basic file with URLs so popups work
if (!file.exists("data/samples_all_wgs.rds")) {
  samples_all_wgs <- st_transform(samples_all, 4326) |>
    mutate(
      gsv_date = NA_character_,
      gsv_maps_url = purrr::map_chr(geometry, ~{
        xy <- st_coordinates(.x)
        paste0("https://www.google.com/maps/@?api=1&map_action=pano&viewpoint=", xy[2], ",", xy[1])
      })
    ) |>
    left_join(st_drop_geometry(sampled_tracts)[, c("tract_id","stratum_id")], by = "tract_id")
  saveRDS(samples_all_wgs, "data/samples_all_wgs.rds")
}

samples_all_wgs   <- readRDS("data/samples_all_wgs.rds")
desired_n <- 3
min_year  <- 2023

# Ensure gsv_year exists
samples_all_wgs <- samples_all_wgs %>%
  mutate(gsv_year = suppressWarnings(as.integer(substr(gsv_date, 1, 4))))

# 1) Up to 3 recent (≥ 2023) per group
recent <- samples_all_wgs %>%
  filter(!is.na(gsv_year) & gsv_year >= min_year) %>%
  group_by(tract_id, source) %>%
  mutate(.rand = runif(dplyr::n())) %>%
  arrange(.by_group = TRUE, .rand) %>%
  slice_head(n = desired_n) %>%
  ungroup() %>%
  select(-.rand)

# 2) Groups that still need filling (make non-sf!)
needs <- samples_all_wgs %>%
  st_drop_geometry() %>%
  group_by(tract_id, source) %>%
  summarise(n_recent = sum(!is.na(gsv_year) & gsv_year >= min_year), .groups = "drop") %>%
  mutate(to_fill = pmax(0, desired_n - n_recent)) %>%
  filter(to_fill > 0)

# 3) Back-fill with most recent < 2023 (or NA)
fallback <- samples_all_wgs %>%
  filter(is.na(gsv_year) | gsv_year < min_year) %>%
  inner_join(needs, by = c("tract_id","source")) %>%
  group_by(tract_id, source) %>%
  arrange(desc(coalesce(gsv_year, -Inf)), .by_group = TRUE) %>%
  mutate(.rk = row_number()) %>%
  filter(.rk <= first(to_fill)) %>%
  ungroup() %>%
  select(-.rk, -n_recent, -to_fill)

# 4) Final set
samples_3each <- bind_rows(recent, fallback)

# Check: ≤ 3 per (tract_id × source)
samples_3each %>% count(tract_id, source) %>% arrange(desc(n))

saveRDS(samples_3each,  "data/samples_3each.rds")

stopifnot(exists("samples_3each"), nrow(samples_3each) > 0)
```

-   **Map B:** Validation points with GSV links and image year.

```{r}
#| echo: false
# ================================================================
# 6. INTERACTIVE MAP (layered so red is always visible)
# ================================================================
tract_union <- st_union(sampled_tracts)
streets_in_sample <- suppressWarnings(st_intersection(streets, tract_union))

# 1) bring gsv_date in
if (!"gsv_date" %in% names(samples_3each)) {
  stopifnot("photo_id" %in% names(samples_3each),
            "photo_id" %in% names(samples_all_wgs))
  samples_3each <- samples_3each %>%
    left_join(
      st_drop_geometry(samples_all_wgs) %>%
        select(photo_id, gsv_date, gsv_maps_url),
      by = "photo_id"
    )
}

# 2) derive year + label
samples_3each <- samples_3each %>%
  mutate(
    gsv_year   = as.integer(str_extract(as.character(gsv_date), "\\b(19|20)\\d{2}\\b")),
    year_label = as.character(gsv_year)
  )

# 1) Reproject to WGS84 for Leaflet
streets_ll <- st_transform(streets_in_sample, 4326)
tracts_ll  <- st_transform(sampled_tracts, 4326)
pts_ll     <- st_transform(samples_3each, 4326)

# 2) Build popup HTML on the full points set (then split)
pts_ll$popup_html <- sprintf(
  "<b>%s</b><br/>Year: %s<br/><a href='%s' target='_blank'>Open GSV</a>",
  htmlEscape(pts_ll$source),
  htmlEscape(ifelse(is.na(pts_ll$year_label), "", pts_ll$year_label)),
  htmlEscape(pts_ll$gsv_maps_url)
)

# 3) Split by source
pts_gen <- pts_ll[pts_ll$source == "osm_general", ]
pts_cyc <- pts_ll[pts_ll$source == "osm_cycleway", ]

# 4) Map
m <- leaflet(options = leafletOptions(zoomControl = TRUE)) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  # streets + tracts
  addPolylines(data = streets_ll, color = "#999999", weight = 1, opacity = 0.6, group = "Streets") %>%
  addPolygons(data = tracts_ll, color = "black", weight = 1, opacity = 0.9,
              fill = TRUE, fillColor = "#000000", fillOpacity = 0.05,
              group = "Sampled tracts") %>%
  # points
  addCircleMarkers(data = pts_gen, radius = 4, stroke = FALSE,
                   fillColor = "red", fillOpacity = 0.6,
                   popup = ~popup_html, group = "General points") %>%
  addCircleMarkers(data = pts_cyc, radius = 4, stroke = FALSE,
                   fillColor = "blue", fillOpacity = 0.6,
                   popup = ~popup_html, group = "Cycleway points") %>%
  # labels (optional)
  addLabelOnlyMarkers(
    data = pts_gen[!is.na(pts_gen$year_label), ],
    label = ~year_label,
    labelOptions = labelOptions(noHide = TRUE, direction = "right", offset = c(6,0),
                                textOnly = TRUE, style = list("font-size" = "12px")),
    group = "Year labels (general)"
  ) %>%
  addLabelOnlyMarkers(
    data = pts_cyc[!is.na(pts_cyc$year_label), ],
    label = ~year_label,
    labelOptions = labelOptions(noHide = TRUE, direction = "right", offset = c(6,0),
                                textOnly = TRUE, style = list("font-size" = "12px")),
    group = "Year labels (cycleway)"
  ) %>%
  addLayersControl(
    overlayGroups = c("Streets", "Sampled tracts",
                      "General points", "Cycleway points",
                      "Year labels (general)", "Year labels (cycleway)"),
    options = layersControlOptions(collapsed = TRUE)
  )
# ^^^ stop the pipe here (NO trailing %>%)

# Auto-zoom to your points (only if there are any)
if (nrow(pts_ll) > 0) {
  bb <- sf::st_bbox(pts_ll)
  # use [[ ]] or unname() so we pass plain numerics (not named vectors)
  m <- m %>% fitBounds(bb[[1]], bb[[2]], bb[[3]], bb[[4]])
  # equivalently: m <- m %>% fitBounds(unname(bb["xmin"]), unname(bb["ymin"]), unname(bb["xmax"]), unname(bb["ymax"]))
}


m
```

### Main limitations and potential solutions

| Limitation | Problem / Issue | Potential solution | Feasibility |
|-----------------|------------------|---------------------|-----------------|
| **No historical access** – GSV API only returns the latest image. | Prevents systematic multi-year validation (e.g., 2015 or 2019 cannot be checked). | Use manual GSV time slider to select past images. | Accurate but not scalable. |
| **Latest ≠ recent** – the “latest” image may be several years old. | Missing or outdated imagery for the most recent year. | Oversample points/tracts and filter by image year to keep enough valid ones. | Improves coverage but computationally heavy. |

<!-- - Ambiguity at intersections → sometimes unclear if GSV is showing the correct street/direction. -->

```{r}
#| include: false
library(dplyr)
library(sf)

# # ensure year column exists
# samples_all_wgs <- samples_all_wgs %>%
#   mutate(gsv_year = as.integer(substr(as.character(gsv_date), 1, 4)))
# 
# # 1) Which tracts have at least one cycleway point in the oversample?
# cycleway_tracts <- samples_all_wgs %>%
#   filter(source == "osm_cycleway") %>%
#   distinct(tract_id) %>%
#   pull(tract_id)
# 
# # Required general: all sampled tracts
# req_general <- sampled_tracts %>%
#   st_drop_geometry() %>%
#   transmute(tract_id, source = "osm_general")
# 
# # Required cycleway: only those with actual cycleway points
# req_cycle <- sampled_tracts %>%
#   st_drop_geometry() %>%
#   filter(tract_id %in% cycleway_tracts) %>%
#   transmute(tract_id, source = "osm_cycleway")
# 
# # Combine requirements
# req_tbl <- bind_rows(req_general, req_cycle)
# 
# # 2) Count candidate points with recent imagery (≥2023)
# counts_recent <- samples_all_wgs %>%
#   filter(!is.na(gsv_year) & gsv_year >= 2023) %>%
#   count(tract_id, source, name = "n_recent")
# 
# # 3) Compare to the target of 3
# target <- 3L
# coverage_check <- req_tbl %>%
#   left_join(counts_recent, by = c("tract_id", "source")) %>%
#   mutate(n_recent = coalesce(n_recent, 0L),
#          ok = n_recent >= target)
# 
# # 4) Shortfalls
# shortfalls <- coverage_check %>%
#   filter(!ok) %>%
#   arrange(tract_id, source)
# 
# shortfalls
```

```{r}
#| echo: false

# Summarise shortfalls at tract level
# tract_coverage <- coverage_check %>%
#   group_by(tract_id) %>%
#   summarise(ok_all = all(ok), .groups = "drop")
# 
# # Join back to polygons
# tracts_with_cov <- sampled_tracts %>%
#   left_join(tract_coverage, by = "tract_id") %>%
#   mutate(status = ifelse(ok_all, "Complete (3/3)", "Incomplete (<3)"))
# 
# # Map
# tmap_mode("plot")
# tm_shape(sampled_tracts) + 
#   tm_borders(col = "grey60") +
# tm_shape(tracts_with_cov) +
#   tm_fill("status", palette = c("Complete (3/3)" = "lightblue", "Incomplete (<3)" = "tomato"),
#           title = "Coverage status") +
#   tm_borders(lwd = 1, col = "black") +
# tm_layout(legend.outside = TRUE,
#           main.title = "Sample coverage by tract (≥2023 imagery)")
```

```{r}
#| echo: false

# simplify strata into D1..D3 (density), C1..C3 (centrality)
# coverage_grid <- coverage_check %>%
#   left_join(sampled_tracts %>% st_drop_geometry() %>% select(tract_id, stratum_id),
#             by = "tract_id") %>%
#   tidyr::separate(stratum_id, into = c("dens", "cent"), sep = "_") %>%
#   mutate(
#     dens = factor(dens, levels = c("D1","D2","D3")),
#     cent = factor(cent, levels = c("C1","C2","C3"))
#   ) %>%
#   group_by(dens, cent, source) %>%
#   summarise(pct_complete = mean(ok)*100, .groups = "drop")
# 
# ggplot(coverage_grid, aes(x = dens, y = cent, fill = pct_complete)) +
#   geom_tile(color = "white") +
#   geom_text(aes(label = paste0(round(pct_complete), "%")), color = "black") +
#   scale_fill_gradient(low = "tomato", high = "steelblue", name = "% complete") +
#   facet_wrap(~source) +
#   labs(title = "Coverage completeness by stratum (Density × Centrality)",
#        x = "Density stratum", y = "Centrality stratum") +
#   theme_minimal()
```

### Assessment

-   **Multi-year validation** (2015–2019–2023) is **not feasible using the GSV API**, since it does not give historical imagery and the “latest” image may be outdated.

-   **Alternative approaches** (manual time slider, oversampling and filtering) are **possible in theory** but probably too slow and resource-intensive for a multi-city project.

-   **The pilot** might be useful **as a methodological case study**, showing the challenges of validating OSM infrastructure data over time with GSV, publishable?

<!-- ## References -->
