---
# title: "ATRAPA Work Package 1: Built Environment Transformations Dataset"
format: html
editor: visual
bibliography: wp1.bib
---

# Validating OpenStreetMap for Detecting Cycling Infrastructure Change: A Pilot Study in Barcelona

## Introduction

-   **Context**: Urban transformations (e.g. new bike lanes, pedestrian areas) can reshape mobility and health, but studying their effects requires reliable historical data.

-   **Problem**: Standardised datasets of infrastructure change are rarely available across cities and years.

-   **Solution**: Volunteered Geographic Information, especially OpenStreetMap (OSM), provides open and historical data on infrastructure, offering a potential way to track changes over time.

-   **Aim**: Assess how well OSM detects cycling infrastructure (CI) changes (additions/removals), by estimating precision, recall, false negatives, and error-adjusted rates of change.

<!-- - **Outputs**: Measures of precision, recall, false negatives, error-adjusted change estimates, and stratum-level diagnostics. -->

-   **Contribution**: Using Barcelona as a pilot case, this study provides one of the first systematic validations of OSM for detecting changes in cycling infrastructure, introduces an error-adjusted approach to estimate true additions and removals, and delivers city-level diagnostics to inform cross-city comparisons and practical monitoring of urban transformations.

## Data and Methods

### Setting and period

-   **Study area**: Barcelona; period: 2019→2023 (snapshot dates: 2020-01-01 for 2019 and 2024-01-01 for 2023).

### Network data (OSM)

-   **Extraction**: Download OSM linework for both years.

-   **CI selector**: `highway = cycleway`, bicycle-designated paths/tracks, roads with `cycleway:*` tags, `bicycle_road = yes`, and designated `living_street`/`pedestrian`.

-   **Non-CI (general network)**: Base roads excluding all CI tags (strict complement).

-   **Change detection**: Build ADDED (present in 2023, absent 2019) and REMOVED (present 2019, absent 2023) layers.

```{r include=FALSE}
#| label: 01-setup
#| include: false
suppressPackageStartupMessages({
  library(sf)
  library(dplyr)
  library(stringr)
  library(osmdata)
  library(osmextract)
  library(leaflet)
})
```

```{r}
#| label: 02-boundary-build-once
#| eval: false
#| include: false
# Run once to write the perimeter file; keep eval=FALSE if it's already saved.

bb <- getbb("Barcelona, Spain")

od <- opq(bb) |>
  add_osm_feature("boundary", "administrative") |>
  add_osm_feature("admin_level", "8") |>
  add_osm_feature("name", "Barcelona") |>
  osmdata_sf()

perim <- od$osm_multipolygons
if (is.null(perim) || nrow(perim) == 0) perim <- od$osm_polygons
stopifnot(nrow(perim) > 0)

# Prefer municipality (Wikidata Q1492) else largest polygon
if ("wikidata" %in% names(perim) && any(perim$wikidata == "Q1492")) {
  perim <- perim[perim$wikidata == "Q1492", ]
} else {
  perim <- perim[which.max(st_area(perim)), ]
}

perim <- st_make_valid(perim) |> st_cast("MULTIPOLYGON") |> st_set_crs(4326)

st_write(perim, "data/barcelona_perimeter.gpkg",
         layer = "barcelona_perimeter", driver = "GPKG", append = FALSE)
```

```{r}
#| label: 03-download-lines-once
#| eval: false
#| include: false
# Run once per version; keep eval=FALSE if files exist.

perimeter_path <- "data/barcelona_perimeter.gpkg"
stopifnot(file.exists(perimeter_path))
city_perimeter <- sf::st_read(perimeter_path, layer = "barcelona_perimeter", quiet = TRUE)
if (is.na(sf::st_crs(city_perimeter)$epsg) || sf::st_crs(city_perimeter)$epsg != 4326) {
  city_perimeter <- sf::st_transform(city_perimeter, 4326)
}

allowed_highway <- c(
  "living_street","pedestrian","cycleway","motorway","trunk","primary","secondary",
  "tertiary","unclassified","residential","motorway_link","trunk_link","primary_link",
  "secondary_link","tertiary_link","service","track","bus_guideway","escape","raceway","busway"
)

fetch_and_crop_data <- function(version, perimeter, output_file) {
  if (file.exists(output_file)) {
    message("↪ Skipping (exists): ", basename(output_file)); return(invisible(NULL))
  }
  # Always pull from "spain" (works for 2017/2019/2023), read only 'lines'
  lines_data <- tryCatch({
    oe_get(
      place   = "spain",
      version = version,
      layer   = "lines",
      boundary = st_bbox(perimeter), boundary_type = "clipsrc",  # fast pre-clip
      extra_tags = c("cycleway","cycleway:left","cycleway:right","bicycle","route"),
      quiet = FALSE
    )
  }, error = function(e) { message("Error fetching ", version, ": ", e$message); return(NULL) })

  if (is.null(lines_data) || nrow(lines_data) == 0) {
    message("No data returned for version ", version); return(invisible(NULL))
  }

  # Keep only (multi)lines
  is_line <- sf::st_geometry_type(lines_data) %in% c("LINESTRING","MULTILINESTRING")
  lines_data <- lines_data[is_line, ]
  if (nrow(lines_data) == 0) { message("No LINESTRINGs for version ", version); return(invisible(NULL)) }

  # Filter to common road classes (do it in R to avoid GDAL vectortranslate issues)
  if ("highway" %in% names(lines_data)) {
    lines_data <- dplyr::filter(lines_data, highway %in% allowed_highway)
  }

  # Final polygon clip
  perim_valid <- sf::st_make_valid(perimeter)
  lines_data <- lines_data[sf::st_intersects(lines_data, perim_valid, sparse = FALSE), ]
  lines_data <- sf::st_intersection(lines_data, perim_valid)

  out_layer <- sub("\\.gpkg$", "", basename(output_file))
  sf::st_write(lines_data, output_file, layer = out_layer, driver = "GPKG", append = FALSE, quiet = TRUE)
  message("✓ Saved ", basename(output_file), " (", nrow(lines_data), " features)")
}

# Exactly these years (add 2017 if you want by uncommenting)
# fetch_and_crop_data("170101", city_perimeter, "data/barcelona_170101_lines.gpkg")
fetch_and_crop_data("200101", city_perimeter, "data/barcelona_19_lines.gpkg")
fetch_and_crop_data("240101", city_perimeter, "data/barcelona_23_lines.gpkg")

message("Done: Barcelona (files in data/)")
```

```{r}
#| label: 04-setup-io
#| message: false
#| warning: false
#| include: false

library(sf); library(dplyr); library(stringr); library(leaflet)

# ---- params ----
city_tag <- "barcelona"
ver19    <- "19"
ver23    <- "23"
crs_work <- 25831
tol_m    <- 10
min_len  <- 30

# ---- paths ----
proc_dir <- "data"
gpkg19   <- file.path(proc_dir, paste0(city_tag, "_", ver19, "_lines.gpkg"))
gpkg23   <- file.path(proc_dir, paste0(city_tag, "_", ver23, "_lines.gpkg"))
lyr19    <- paste0(city_tag, "_", ver19, "_lines")
lyr23    <- paste0(city_tag, "_", ver23, "_lines")

stopifnot(file.exists(gpkg19), file.exists(gpkg23))

# ---- read ----
l19 <- st_read(gpkg19, layer = lyr19, quiet = TRUE)
l23 <- st_read(gpkg23, layer = lyr23, quiet = TRUE)
```

```{r}
#| label: 05-helpers
#| message: false
#| warning: false
#| include: false

# Lowercase-safe accessors
.get_chr <- function(x, col){
  nm <- names(x); n <- nrow(x)
  if (col %in% nm) tolower(as.character(x[[col]])) else rep(NA_character_, n)
}

# STRICT cycle infra selector (clear physical infra + designated paths + bicycle road)
pick_cycle_strict <- function(x){
  stopifnot(inherits(x,"sf")); if (!nrow(x)) return(x)
  highway      <- .get_chr(x, "highway")
  bicycle      <- .get_chr(x, "bicycle")
  bicycle_road <- .get_chr(x, "bicycle_road")

  lane_cols <- c("cycleway","cycleway:both","cycleway:left","cycleway:right")
  lane_vals <- c("lane","track","opposite_lane","opposite_track","separate")

  n <- nrow(x); has_lane <- rep(FALSE, n)
  for (col in lane_cols) if (col %in% names(x)) {
    v <- .get_chr(x, col)
    has_lane <- has_lane | (!is.na(v) & v %in% lane_vals)
  }

  is_cycleway        <- !is.na(highway) & highway == "cycleway"
  is_designated_path <- (!is.na(highway) & highway %in% c("path","footway","track")) &
                        (!is.na(bicycle) & bicycle %in% c("designated","yes"))
  is_bicycle_road    <- !is.na(bicycle_road) & bicycle_road == "yes"

  keep <- is_cycleway | has_lane | is_designated_path | is_bicycle_road
  x[keep, , drop = FALSE]
}

# STRICT non-CI (NA-safe complement among base road classes)
pick_noncycle_strict <- function(x){
  stopifnot(inherits(x,"sf")); if (!nrow(x)) return(x)
  highway      <- .get_chr(x, "highway")
  bicycle      <- .get_chr(x, "bicycle")
  bicycle_road <- .get_chr(x, "bicycle_road")

  lane_cols <- c("cycleway","cycleway:both","cycleway:left","cycleway:right")
  lane_vals <- c("lane","track","opposite_lane","opposite_track","separate")

  n <- nrow(x); has_lane <- rep(FALSE, n)
  for (col in lane_cols) if (col %in% names(x)) {
    v <- .get_chr(x, col)
    has_lane <- has_lane | (!is.na(v) & v %in% lane_vals)
  }

  is_cycleway        <- !is.na(highway) & highway == "cycleway"
  is_designated_path <- (!is.na(highway) & highway %in% c("path","footway","track")) &
                        (!is.na(bicycle) & bicycle %in% c("designated","yes"))
  is_bicycle_road    <- !is.na(bicycle_road) & bicycle_road == "yes"

  # CI definition mirrors pick_cycle_strict
  is_ci <- is_cycleway | has_lane | is_designated_path | is_bicycle_road

  base_ok <- !is.na(highway) & highway %in% c(
    "trunk","primary","secondary","tertiary",
    "unclassified","residential","service",
    "primary_link","secondary_link","tertiary_link",
    "living_street","pedestrian"
  )

  x[ base_ok & !is_ci, , drop = FALSE]
}

# Normalize to single LINESTRINGs, valid geometries
normalize_geoms <- function(x){
  if (!nrow(x)) return(x)
  x <- st_make_valid(x)
  x <- suppressWarnings(st_cast(x, "MULTILINESTRING"))
  x <- st_line_merge(x)
  g <- st_geometry(x)
  g <- suppressWarnings(st_cast(g, "LINESTRING"))
  st_set_geometry(x, g) |> st_make_valid()
}

# Length filter (meters)
len_ok <- function(x, min_len = 0){
  if (!inherits(x,"sf") || !nrow(x) || min_len <= 0) return(x)
  L <- as.numeric(st_length(x))
  x[L >= min_len, , drop = FALSE]
}

```

```{r}
#| label: 06-build-ci-change
#| message: false
#| warning: false
# Build CI networks (2019/2023) + Added/Removed

# Cycle infra by tags
cyc19 <- l19 |> pick_cycle_strict()
cyc23 <- l23 |> pick_cycle_strict()

# Work CRS + normalize
cyc19_n <- cyc19 |> st_transform(crs_work) |> normalize_geoms()
cyc23_n <- cyc23 |> st_transform(crs_work) |> normalize_geoms()

# Buffers for change detection
buf19 <- st_buffer(st_geometry(cyc19_n), tol_m) |> st_union()
buf23 <- st_buffer(st_geometry(cyc23_n), tol_m) |> st_union()

# Differences (element-wise vs union buffer)
added_raw   <- st_difference(st_geometry(cyc23_n), buf19)
removed_raw <- st_difference(st_geometry(cyc19_n), buf23)

# Extract clean LINESTRINGs
process_difference <- function(diff_geom, crs){
  if (length(diff_geom) == 0) return(st_sf(geometry = st_sfc(crs = crs)))
  g <- suppressWarnings(st_collection_extract(diff_geom, "LINESTRING"))
  g <- suppressWarnings(st_cast(g, "LINESTRING"))
  out <- st_sf(geometry = g, crs = crs)
  out <- out[as.numeric(st_length(out)) > 0, , drop = FALSE]
  out
}

added   <- process_difference(added_raw,   crs_work) |> len_ok(min_len)
removed <- process_difference(removed_raw, crs_work) |> len_ok(min_len)

message("CI nets — 2019: ", nrow(cyc19_n), " | 2023: ", nrow(cyc23_n),
        " | Added: ", nrow(added), " | Removed: ", nrow(removed))
```

```{r}
#| label: nonci-network-2023
#| include: false
#| message: false
#| warning: false

# If cyc23_n isn't available, build it quickly
if (!exists("cyc23_n")) {
  cyc23 <- l23 |>
    pick_cycle_strict() |>
    sf::st_transform(crs_work) |>
    sf::st_make_valid()

  cyc23_n <- cyc23 |>
    suppressWarnings(sf::st_cast("MULTILINESTRING")) |>
    sf::st_line_merge() |>
    suppressWarnings(sf::st_cast("LINESTRING"))
}

# Base roads to consider "general street network"
base_roads <- c(
  "trunk","primary","secondary","tertiary",
  "unclassified","residential","service",
  "primary_link","secondary_link","tertiary_link",
  "living_street","pedestrian"
)

# 2023 base network (lines only, in work CRS)
base23 <- l23 |>
  sf::st_transform(crs_work) |>
  dplyr::filter(.data$highway %in% base_roads)

keep_lines <- sf::st_geometry_type(base23) %in% c("LINESTRING","MULTILINESTRING")
base23 <- base23[keep_lines, , drop = FALSE] |>
  sf::st_make_valid() |>
  suppressWarnings(sf::st_cast("LINESTRING"))

# Subtract a small buffer around 2023 CI network
cyc_buf <- sf::st_buffer(sf::st_geometry(cyc23_n), tol_m) |> sf::st_union()

diff_geom <- sf::st_difference(sf::st_geometry(base23), cyc_buf)
general23_n <- sf::st_collection_extract(diff_geom, "LINESTRING", warn = FALSE) |>
  sf::st_sf(crs = crs_work)

# Clean: drop empties + (optional) length filter
general23_n <- general23_n[!sf::st_is_empty(general23_n), , drop = FALSE]
if (exists("min_len")) {
  general23_n <- general23_n[as.numeric(sf::st_length(general23_n)) >= min_len, , drop = FALSE]
}
```

```{r}
#| label: 08-read-boundary-for-map
#| message: false
#| warning: false
# Read perimeter (WGS84) for later map display / fitBounds

perimeter_path <- "data/barcelona_perimeter.gpkg"
stopifnot(file.exists(perimeter_path))
city_perimeter <- sf::st_read(perimeter_path, layer = "barcelona_perimeter", quiet = TRUE)
city_perimeter <- sf::st_transform(city_perimeter, 4326)

# (Stop here — next chunk will be the map)
```

```{r}
#| label: 09-map
#| echo: false
#| message: false
#| warning: false

# ---- helper: force WGS84 LINESTRING and skip empties -------------------------
as_wgs_lines <- function(x){
  if (!inherits(x, "sf") || !nrow(x)) return(NULL)
  x <- sf::st_make_valid(x)
  keep <- sf::st_geometry_type(x) %in% c("LINESTRING","MULTILINESTRING")
  x <- x[keep, , drop = FALSE]
  if (!nrow(x)) return(NULL)
  x <- suppressWarnings(sf::st_cast(x, "LINESTRING"))
  x <- x[!sf::st_is_empty(sf::st_geometry(x)), , drop = FALSE]
  if (!nrow(x)) return(NULL)
  x <- sf::st_transform(x, 4326)
  x
}

as_wgs_poly <- function(x){
  if (!inherits(x, "sf") || !nrow(x)) return(NULL)
  x <- sf::st_make_valid(x)
  keep <- sf::st_geometry_type(x) %in% c("POLYGON","MULTIPOLYGON")
  x <- x[keep, , drop = FALSE]
  if (!nrow(x)) return(NULL)
  x <- sf::st_transform(x, 4326)
  x
}

# ---- colours -----------------------------------------------------------------
col_2019 <- "#666666"
col_added <- "#D95F02"
col_removed <- "#FFFFFF"
col_removed_halo <- "#000000"
col_2023 <- "#1B9E77"
col_nonci <- "#BDBDBD"

alpha_2019  <- 0.85
alpha_added <- 0.95
alpha_2023  <- 0.65
alpha_nonci <- 0.70
alpha_halo  <- 0.18

# ---- prep layers safely ------------------------------------------------------
perim_wgs   <- if (exists("city_perimeter")) as_wgs_poly(city_perimeter) else NULL
cyc19_wgs   <- if (exists("cyc19_n"))  as_wgs_lines(cyc19_n)  else NULL
added_wgs   <- if (exists("added"))    as_wgs_lines(added)    else NULL
removed_wgs <- if (exists("removed"))  as_wgs_lines(removed)  else NULL
cyc23_wgs   <- if (exists("cyc23_n"))  as_wgs_lines(cyc23_n)  else NULL
nonci_wgs   <- if (exists("general23_n")) as_wgs_lines(general23_n) else NULL

# ---- bounds (prefer perimeter, else any layer, else BCN-ish fallback) --------
get_bbox <- function(obj) if (!is.null(obj) && inherits(obj,"sf") && nrow(obj)) sf::st_bbox(obj) else NULL
bb <- get_bbox(perim_wgs)
if (is.null(bb)) {
  cands <- Filter(Negate(is.null), list(get_bbox(cyc19_wgs), get_bbox(added_wgs),
                                        get_bbox(removed_wgs), get_bbox(cyc23_wgs),
                                        get_bbox(nonci_wgs)))
  if (length(cands)) {
    mins <- do.call(pmin, lapply(cands, function(b) c(b$xmin,b$ymin)))
    maxs <- do.call(pmax, lapply(cands, function(b) c(b$xmax,b$ymax)))
    bb <- structure(list(xmin=mins[1], ymin=mins[2], xmax=maxs[1], ymax=maxs[2]), class="bbox")
  } else {
    bb <- c(xmin = 2.05, ymin = 41.30, xmax = 2.25, ymax = 41.45)
  }
}
bounds <- unname(c(bb["xmin"], bb["ymin"], bb["xmax"], bb["ymax"]))

# ---- map ---------------------------------------------------------------------
m <- leaflet(options = leafletOptions(preferCanvas = TRUE)) %>%
  addProviderTiles("CartoDB.Positron", group = "Positron")

if (!is.null(perim_wgs))
  m <- m %>% addPolygons(data = perim_wgs, group = "Boundary",
                         weight = 2, color = "#222222", fill = FALSE, fillOpacity = 0)

if (!is.null(cyc19_wgs))
  m <- m %>% addPolylines(data = cyc19_wgs, group = "2019 CI",
                          weight = 2, color = col_2019, opacity = alpha_2019)

if (!is.null(removed_wgs)) {
  m <- m %>% addPolylines(data = removed_wgs, group = "Removed (19→23)",
                          weight = 2.6, color = col_removed_halo, opacity = alpha_halo)
  m <- m %>% addPolylines(data = removed_wgs, group = "Removed (19→23)",
                          weight = 2.0, color = col_removed, opacity = 1.0)
}

if (!is.null(added_wgs))
  m <- m %>% addPolylines(data = added_wgs, group = "Added (19→23)",
                          weight = 2, color = col_added, opacity = alpha_added)

if (!is.null(cyc23_wgs))
  m <- m %>% addPolylines(data = cyc23_wgs, group = "2023 CI",
                          weight = 2, color = col_2023, opacity = alpha_2023)

if (!is.null(nonci_wgs))
  m <- m %>% addPolylines(data = nonci_wgs, group = "General (non-CI) 2023",
                          weight = 1, color = col_nonci, opacity = alpha_nonci,
                          smoothFactor = 0.5)

m %>%
  addLayersControl(
    baseGroups = c("Positron"),
    overlayGroups = c("Boundary","2019 CI","Added (19→23)","Removed (19→23)","2023 CI","General (non-CI) 2023"),
    options = layersControlOptions(collapsed = TRUE)
  ) %>%
  hideGroup(c("2023 CI","General (non-CI) 2023")) %>%
  fitBounds(bounds[1], bounds[2], bounds[3], bounds[4]) %>%
  addLegend(
    position = "bottomright", title = "Layers",
    colors = c(col_2019, col_added, col_removed, col_2023, col_nonci),
    labels = c("2019 CI","Added (19→23)","Removed (19→23)","2023 CI","General (non-CI)"),
    opacity = 1
  )
```


### Sampling frame and design

-   **Stratified tracts**: Pre-selected census tracts by centrality × density (3×3).

-   **Per-tract quotas (example)**: ADD=2, REMOVE=1, GENERAL=2 (GENERAL = 2023 non-CI network). Length-weighted sampling within tract; one midpoint per sampled segment.

-   **Expected N**: At most #tracts × (2+1+2), reduced where tracts have fewer or no candidates in one or more categories (ADD, REMOVE, GENERAL).

```{r}
#| include: false
# ================================================================
# STRATIFIED SAMPLING AND STREET VIEW LINK GENERATION
# ================================================================

# ---- Load required libraries ----
library(sf)
library(dplyr)
library(tidyr)
library(biscale)
library(ggplot2)
library(cowplot)
library(osmdata)
library(purrr)
library(jsonlite)
library(httr)
library(tmap)
library(stringr)
library(sf)
library(leaflet)
library(htmltools)
```

```{r}
#| include: false
# ================================================================
# 1. LOAD AND PREPARE DATA
# ================================================================

# Load spatial and population data
tracts_geo <- readRDS("data/bcn-seccio-censal/BCN_seccion_censal_2015.rds")
tracts_pop <- readRDS("data/pop/Population_Census_tract_2015_2022.rds")

tracts_pop_wide <- tracts_pop |> 
  filter(Year %in% c("2015", "2019", "2022")) |> 
  pivot_wider(names_from = Year, values_from = Population, names_prefix = "pop_")

# Merge geometries with population data
barcelona_tracts <- left_join(tracts_geo, tracts_pop_wide, by = c("CUSEC" = "GEOID")) |> 
  st_transform(25831)

# Compute area, population density, and distance to city center
placa_catalunya <- st_sfc(st_point(c(431870, 4581450)), crs = 25831)

barcelona_tracts <- barcelona_tracts |> 
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    dens_2015 = pop_2015 / area_km2,
    dens_2019 = pop_2019 / area_km2,
    dens_2022 = pop_2022 / area_km2,
    centroid = st_centroid(geometry),
    dist_centre_km = as.numeric(st_distance(centroid, placa_catalunya)) / 1000
  )
```

```{r}
#| message: false
#| warning: false
#| include: false
#| paged-print: false

# ================================================================
# 2. STRATIFIED SAMPLING
# ================================================================

barcelona_tracts <- barcelona_tracts |>
  mutate(
    dens_stratum = ntile(dens_2022, 3),
    cent_stratum = ntile(-dist_centre_km, 3),
    stratum_id = paste0("D", dens_stratum, "_C", cent_stratum)
  )

set.seed(123)
per_stratum <- 5  # try 3–6 depending on effort

sampled_tracts <- barcelona_tracts |> 
  group_by(stratum_id) |> 
  sample_n(size = min(per_stratum, n())) |> 
  ungroup()
```

```{r}
#| echo: false

# ================================================================
# 3. BIVARIATE MAPPING
# ================================================================

barcelona_tracts <- barcelona_tracts |> 
mutate(centrality_flipped = -dist_centre_km)
 
bb_bivar <- bi_class(barcelona_tracts, x = dens_2022, y = centrality_flipped, style = "quantile", dim = 3)
 
bivar_map <- ggplot() +
   geom_sf(data = bb_bivar, aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
   bi_scale_fill(pal = "DkBlue2", dim = 3) +
   geom_sf(data = sampled_tracts, fill = NA, color = "black", linewidth = 0.5) +
   labs(title = "") +
   bi_theme() +
   theme(plot.title = element_text(size = 12, hjust = 0.5))
 
 bivar_legend <- bi_legend(pal = "DkBlue2", dim = 3, xlab = "Higher Density", ylab = "More Central", size = 9) +
   theme(axis.title = element_text(size = 9), axis.text = element_blank(),
         plot.margin = margin(5, 5, 5, 5), panel.background = element_blank(), plot.background = element_blank())
 
 final_plot <- ggdraw() +
   draw_plot(bivar_map, 0, 0, 1, 1) +
   draw_plot(bivar_legend, x = 0.73, y = 0.03, width = 0.26, height = 0.26)
 
 final_plot
 
 ggsave("figs/stratified_sample_bivariate_map.png", final_plot, width = 8, height = 8, dpi = 300)
```

### Validation source and coding

-   **Imagery**: Google Street View (GSV).

-   **Tolerance windows**: 2019 target: ±1 year; 2023 target: ±2 years (fixed across coders).

-   **Per-point fields (Excel)**:

    -   *Metadata*: class (ADD/REMOVE/GENERAL), tract_id, stratum, lon, lat, gsv_link (+ clickable “Open GSV”).

    -   *Verifiability*: verifiable_2019, verifiable_2023 (TRUE/FALSE).

    -   *Presence*: present_2019, present_2023 (TRUE/FALSE/NA).

    -   *Notes*: (obstruction, roadworks, poor angle, etc.).

-   **Consistency rules**:

    -   If verifiable\_\* = FALSE → set corresponding present\_\* = NA.

    -   If imagery exists but unclear → keep verifiable\_\* = TRUE, set present\_\* = NA.

```{r}
#| eval: false
#| include: false
# ======================= CLEAN VALIDATION PIPELINE (robust) ===================
library(sf)
library(dplyr)

# --- small guard: keep only LINESTRINGs ---------------------------------------
ensure_lines <- function(x){
  if (!inherits(x,"sf") || !nrow(x)) return(x[0, , drop = FALSE])
  g <- sf::st_geometry(x)
  g <- sf::st_collection_extract(g, "LINESTRING", warn = FALSE)
  g <- suppressWarnings(sf::st_cast(g, "LINESTRING"))
  keep <- !sf::st_is_empty(g)
  sf::st_sf(sf::st_drop_geometry(x)[keep, , drop = FALSE], geometry = g[keep], crs = sf::st_crs(x))
}

# --- 0) Helpers ---------------------------------------------------------------
# Midpoint per line (robust to MULTI/collections; assumes LINESTRING input)
midpoint_one <- function(x){
  if (!inherits(x,"sf") || nrow(x) == 0)
    return(sf::st_sf(id = integer(0), geometry = sf::st_sfc(crs = sf::st_crs(x))))
  lx <- ensure_lines(sf::st_make_valid(x))
  if (!nrow(lx)) return(lx[0, , drop = FALSE])
  pts <- sf::st_line_sample(sf::st_geometry(lx), sample = 0.5)
  sf::st_as_sf(
    data.frame(id = seq_along(pts)),
    geometry = sf::st_cast(pts, "POINT"),
    crs = sf::st_crs(lx)
  )
}

# Length-weighted row indices (safe on zero/NA lengths)
w_sample_idx <- function(x, n){
  if (!inherits(x,"sf") || nrow(x) == 0 || n <= 0) return(integer(0))
  L <- as.numeric(sf::st_length(x)); S <- sum(L, na.rm = TRUE)
  if (!is.finite(S) || S == 0) {
    sample(seq_len(nrow(x)), size = min(n, nrow(x)), replace = FALSE)
  } else {
    sample(seq_len(nrow(x)), size = min(n, nrow(x)), replace = FALSE, prob = L/S)
  }
}

# Points → lon/lat + GSV link (0-row safe; preserves existing columns)
to_lonlat_tbl <- function(pts_sf){
  stopifnot(inherits(pts_sf, "sf"))
  n <- nrow(pts_sf)
  if (n == 0){
    base <- sf::st_drop_geometry(pts_sf)[0, , drop = FALSE]
    base$lon <- numeric(0); base$lat <- numeric(0); base$gsv_link <- character(0)
    return(base)
  }
  wgs    <- sf::st_transform(pts_sf, 4326)
  coords <- sf::st_coordinates(wgs)
  out    <- sf::st_drop_geometry(wgs)
  out$lon <- coords[,1]; out$lat <- coords[,2]
  out$gsv_link <- paste0("https://www.google.com/maps/@?api=1&map_action=pano&viewpoint=",
                         out$lat, ",", out$lon)
  out
}

# Add validation placeholders (NA logical flags + notes) — 0-row safe
add_validation_cols <- function(df){
  n <- nrow(df)
  df$present_2019    <- rep(NA, n)
  df$present_2023    <- rep(NA, n)
  df$verifiable_2019 <- rep(NA, n)
  df$verifiable_2023 <- rep(NA, n)
  df$notes           <- rep(NA_character_, n)
  df
}

# Coerce critical columns so bind_rows never complains
coerce_for_bind <- function(df){
  if (!"class" %in% names(df))    df$class <- character(nrow(df))
  if (!"interval" %in% names(df)) df$interval <- character(nrow(df))
  df$class            <- as.character(df$class)
  df$interval         <- as.character(df$interval)
  if ("gsv_link" %in% names(df)) df$gsv_link <- as.character(df$gsv_link)
  if ("lon" %in% names(df))      df$lon      <- as.numeric(df$lon)
  if ("lat" %in% names(df))      df$lat      <- as.numeric(df$lat)
  if ("present_2019" %in% names(df))    df$present_2019    <- as.logical(df$present_2019)
  if ("present_2023" %in% names(df))    df$present_2023    <- as.logical(df$present_2023)
  if ("verifiable_2019" %in% names(df)) df$verifiable_2019 <- as.logical(df$verifiable_2019)
  if ("verifiable_2023" %in% names(df)) df$verifiable_2023 <- as.logical(df$verifiable_2023)
  if ("notes" %in% names(df))           df$notes           <- as.character(df$notes)
  df
}

# --- 1) Sample sizes ----------------------------------------------------------
set.seed(42)
N_ADD <- min(120, nrow(added))
N_REM <- min(60,  nrow(removed))
N_GEN <- 120

# --- 2) ADDED / REMOVED → midpoints ------------------------------------------
added_lns   <- ensure_lines(added)
removed_lns <- ensure_lines(removed)

add_idx <- w_sample_idx(added_lns,   N_ADD)
rem_idx <- w_sample_idx(removed_lns, N_REM)

added_pts   <- midpoint_one(added_lns[add_idx, ])   |> dplyr::mutate(class = "ADD")
removed_pts <- midpoint_one(removed_lns[rem_idx, ]) |> dplyr::mutate(class = "REMOVE")

# --- 3) GENERAL (2023 non-cycle) → midpoints ---------------------------------
# Prefer your prebuilt non-CI network if available; otherwise tag-based fallback
if (exists("general23_n") && inherits(general23_n, "sf") && nrow(general23_n)) {
  noncycle23 <- ensure_lines(sf::st_transform(general23_n, crs_work))
} else {
  base23 <- l23 |>
    sf::st_transform(crs_work)

  # keep only LINESTRING / MULTILINESTRING before tag logic
  keep_lines <- sf::st_geometry_type(base23, by_geometry = TRUE) %in% c("LINESTRING","MULTILINESTRING")
  base23 <- base23[keep_lines, , drop = FALSE]

  noncycle23 <- base23 |>
    pick_noncycle_strict() |>
    sf::st_make_valid() |>
    ensure_lines()
}

# Drop empties
if (nrow(noncycle23))
  noncycle23 <- noncycle23[!sf::st_is_empty(sf::st_geometry(noncycle23)), , drop = FALSE]

gen_idx <- w_sample_idx(noncycle23, N_GEN)
gen_pts <- midpoint_one(noncycle23[gen_idx, ]) |> dplyr::mutate(class = "GENERAL")

# --- 4) Tables with placeholders, bind ----------------------------------------
added_tbl   <- to_lonlat_tbl(added_pts)   |> dplyr::mutate(interval = "2019→2023") |> add_validation_cols() |> coerce_for_bind()
removed_tbl <- to_lonlat_tbl(removed_pts) |> dplyr::mutate(interval = "2019→2023") |> add_validation_cols() |> coerce_for_bind()
general_tbl <- to_lonlat_tbl(gen_pts)     |> dplyr::mutate(interval = "2019→2023") |> add_validation_cols() |> coerce_for_bind()

samples_tbl <- dplyr::bind_rows(added_tbl, removed_tbl, general_tbl)

# --- (optional) sanity print --------------------------------------------------
cat("rows → ADD:", nrow(added_tbl), " REMOVE:", nrow(removed_tbl), " GENERAL:", nrow(general_tbl), "\n")
# readr::write_csv(samples_tbl, "samples_2019_2023.csv")
# ===================== END CLEAN VALIDATION PIPELINE ==========================


# 5) Write Excel (single sheet)
suppressPackageStartupMessages(requireNamespace("writexl", quietly = TRUE))
library(openxlsx); library(dplyr)

outfile <- "samples_2019_2023.xlsx"

df_xlsx <- samples_tbl %>%
  mutate(GSV = NA_character_) %>%
  relocate(GSV, .after = gsv_link)

wb <- createWorkbook()
addWorksheet(wb, "validation_samples")
writeData(wb, "validation_samples", df_xlsx)

n <- nrow(samples_tbl)
if (n > 0) {
  link_col <- which(names(df_xlsx) == "GSV")
  formulas <- paste0('HYPERLINK("', samples_tbl$gsv_link, '","Open GSV")')
  writeFormula(wb, sheet = "validation_samples",
               x = formulas, startCol = link_col, startRow = 2)
}
addFilter(wb, "validation_samples", row = 1, cols = 1:ncol(df_xlsx))
setColWidths(wb, "validation_samples", cols = 1:ncol(df_xlsx), widths = "auto")
freezePane(wb, "validation_samples", firstActiveRow = 2, firstActiveCol = 1)
saveWorkbook(wb, outfile, overwrite = TRUE)
message("Wrote: ", normalizePath(outfile))

# (Optional) write sampled points to GPKG
out_gpkg <- file.path("data","validation_2019_2023_points.gpkg")
if (nrow(added_pts))   sf::st_write(added_pts,   out_gpkg, "Added_pts",   delete_layer = TRUE, quiet = TRUE)
if (nrow(removed_pts)) sf::st_write(removed_pts, out_gpkg, "Removed_pts", delete_layer = TRUE, quiet = TRUE)
if (nrow(gen_pts))     sf::st_write(gen_pts,     out_gpkg, "General_pts", delete_layer = TRUE, quiet = TRUE)
cat("Wrote:", out_gpkg, "\n")

```

```{r}
#| label: helpers-validation
#| echo: false
#| message: false
#| warning: false
#| eval: true
#| include: false

library(sf); library(dplyr)

# 1) One-point-per-row midpoint (no row explosion)
if (!exists("midpoint_one")) {
  midpoint_one <- function(x){
    if (!inherits(x,"sf") || !nrow(x))
      return(sf::st_sf(id=integer(0), geometry=sf::st_sfc(crs=sf::st_crs(x))))
    g <- sf::st_geometry(x) |> sf::st_make_valid() |> suppressWarnings(sf::st_line_merge())
    sf::st_as_sf(data.frame(id=seq_len(nrow(x))),
                 geometry = sf::st_point_on_surface(g),
                 crs = sf::st_crs(x))
  }
}

# 2) Points -> lon/lat + GSV
if (!exists("to_lonlat_tbl")) {
  to_lonlat_tbl <- function(pts_sf){
    stopifnot(inherits(pts_sf,"sf"))
    n <- nrow(pts_sf)
    if (!n) {
      base <- sf::st_drop_geometry(pts_sf)[0,,drop=FALSE]
      base$lon <- numeric(0); base$lat <- numeric(0); base$gsv_link <- character(0)
      return(base)
    }
    wgs <- sf::st_transform(pts_sf, 4326); xy <- sf::st_coordinates(wgs)
    out <- sf::st_drop_geometry(wgs)
    out$lon <- xy[,1]; out$lat <- xy[,2]
    out$gsv_link <- paste0("https://www.google.com/maps/@?api=1&map_action=pano&viewpoint=", out$lat, ",", out$lon)
    out
  }
}

# 3) Validation columns
if (!exists("add_validation_cols")) {
  add_validation_cols <- function(df){
    n <- nrow(df)
    df$present_2019    <- rep(NA, n)
    df$present_2023    <- rep(NA, n)
    df$verifiable_2019 <- rep(NA, n)
    df$verifiable_2023 <- rep(NA, n)
    df$notes           <- rep(NA_character_, n)
    df
  }
}

# 4) Coerce for bind_rows
if (!exists("coerce_for_bind")) {
  coerce_for_bind <- function(df){
    if (!"class" %in% names(df))    df$class <- character(nrow(df))
    if (!"interval" %in% names(df)) df$interval <- character(nrow(df))
    df$class    <- as.character(df$class)
    df$interval <- as.character(df$interval)
    if ("gsv_link" %in% names(df)) df$gsv_link <- as.character(df$gsv_link)
    if ("lon" %in% names(df))      df$lon      <- as.numeric(df$lon)
    if ("lat" %in% names(df))      df$lat      <- as.numeric(df$lat)
    if ("present_2019" %in% names(df))    df$present_2019    <- as.logical(df$present_2019)
    if ("present_2023" %in% names(df))    df$present_2023    <- as.logical(df$present_2023)
    if ("verifiable_2019" %in% names(df)) df$verifiable_2019 <- as.logical(df$verifiable_2019)
    if ("verifiable_2023" %in% names(df)) df$verifiable_2023 <- as.logical(df$verifiable_2023)
    if ("notes" %in% names(df))           df$notes           <- as.character(df$notes)
    df
  }
}

# 5) Non-cycle strict (only real infra counts as cycle)
if (!exists("pick_noncycle_strict")) {
  pick_noncycle_strict <- function(x){
    stopifnot(inherits(x,"sf")); if (!nrow(x)) return(x)
    nm <- names(x); n <- nrow(x)
    gchr <- function(col) if (col %in% nm) tolower(as.character(x[[col]])) else rep(NA_character_, n)
    highway      <- gchr("highway")
    bicycle      <- gchr("bicycle")
    bicycle_road <- gchr("bicycle_road")
    lane_cols <- c("cycleway","cycleway:both","cycleway:left","cycleway:right")
    lane_vals <- c("lane","track","opposite_lane","opposite_track","separate")
    has_lane <- rep(FALSE, n)
    for (col in lane_cols) if (col %in% nm) { v <- gchr(col); has_lane <- has_lane | (v %in% lane_vals) }
    is_cycleway <- highway == "cycleway"
    is_path     <- (highway %in% c("path","footway","track")) & (bicycle %in% c("designated","yes"))
    is_bike_rd  <- bicycle_road == "yes"
    keep_cycle  <- is_cycleway | has_lane | is_path | is_bike_rd
    base_ok <- highway %in% c("trunk","primary","secondary","tertiary",
                              "unclassified","residential","service",
                              "primary_link","secondary_link","tertiary_link",
                              "living_street","pedestrian")
    x[base_ok & !keep_cycle, , drop = FALSE]
  }
}
```

```{r}
#| label: validation-stratified
#| echo: false
#| message: false
#| warning: false
#| eval: true
#| include: false

# safety: define if missing
if (!exists("pick_noncycle_strict")) {
  pick_noncycle_strict <- function(x){
    stopifnot(inherits(x,"sf"))
    if (!nrow(x)) return(x)
    nm <- names(x); n <- nrow(x)
    gchr <- function(col) if (col %in% nm) tolower(as.character(x[[col]])) else rep(NA_character_, n)
    highway      <- gchr("highway")
    bicycle      <- gchr("bicycle")
    bicycle_road <- gchr("bicycle_road")
    lane_cols <- c("cycleway","cycleway:both","cycleway:left","cycleway:right")
    lane_vals <- c("lane","track","opposite_lane","opposite_track","separate")
    has_physical_lane <- rep(FALSE, n)
    for (col in lane_cols) if (col %in% nm) {
      v <- gchr(col); has_physical_lane <- has_physical_lane | (v %in% lane_vals)
    }
    is_cycleway        <- highway == "cycleway"
    is_designated_path <- (highway %in% c("path","footway","track")) & (bicycle %in% c("designated","yes"))
    is_bicycle_road    <- bicycle_road == "yes"
    keep_cycle <- is_cycleway | has_physical_lane | is_designated_path | is_bicycle_road
    base_roads <- highway %in% c("trunk","primary","secondary","tertiary",
                                 "unclassified","residential","service",
                                 "primary_link","secondary_link","tertiary_link",
                                 "living_street","pedestrian")
    x[base_roads & !keep_cycle, , drop = FALSE]
  }
}

# Requires: added, removed, l23, crs_work, sampled_tracts (sf POLYGONS)
stopifnot(exists("added"), exists("removed"), exists("l23"), exists("crs_work"))
stopifnot(exists("sampled_tracts"))

# --- Ensure sampled_tracts has a 'tract_id' ----------------------------------
normalize_tract_id <- function(s){
  nm  <- names(s); nml <- tolower(nm)
  candidates <- c("tract_id","tractid","geoid","geoid10","geoid20","id","code","name","objectid","fid")
  hit <- which(nml %in% candidates)
  if (length(hit) > 0) {
    s$tract_id <- as.character(s[[ nm[hit[1]] ]])
  } else {
    s$tract_id <- sprintf("tract_%04d", seq_len(nrow(s)))
  }
  s
}
sampled_tracts <- normalize_tract_id(sampled_tracts)
stopifnot("tract_id" %in% names(sampled_tracts))

# --- Quotas per tract (tune as needed) ---------------------------------------
PERTRACT_ADD <- 2
PERTRACT_REM <- 1
PERTRACT_GEN <- 2
FORCE_QUOTA  <- FALSE   # TRUE = allow replacement to hit quota

# --- Ensure tracts are in work CRS and have 'stratum' ------------------------
tracts <- sf::st_transform(sampled_tracts, crs_work)
if (!"stratum" %in% names(tracts)) {
  if (all(c("density_f","centrality_f") %in% names(tracts))) {
    tracts$stratum <- interaction(tracts$density_f, tracts$centrality_f, sep = "×", drop = TRUE)
  } else {
    tracts$stratum <- factor("all")
  }
}
tracts$stratum <- droplevels(as.factor(tracts$stratum))

# --- Helper: sample lines within each tract (length-weighted) ----------------
sample_lines_by_tract <- function(lines_sf, tracts_sf, n_per, replace = FALSE, min_len = 0){
  if (!inherits(lines_sf,"sf") || nrow(lines_sf) == 0) {
    out <- lines_sf[0, , drop = FALSE]
    out$tract_id <- character(0); out$stratum <- factor()[0]
    return(out)
  }
  L <- sf::st_transform(lines_sf, sf::st_crs(tracts_sf))
  if (min_len > 0) {
    L <- L[as.numeric(sf::st_length(L)) >= min_len, , drop = FALSE]
  }
  if (!nrow(L)) {
    out <- L; out$tract_id <- character(0); out$stratum <- factor()[0]; return(out)
  }

  hits  <- sf::st_intersects(tracts_sf, L)  # per-tract list of row indices
  picks <- vector("list", length = nrow(tracts_sf))

  for (i in seq_len(nrow(tracts_sf))) {
    idx <- hits[[i]]
    if (!length(idx)) next

    draw_with_replacement <- isTRUE(replace) && length(idx) < n_per
    k <- if (draw_with_replacement) n_per else min(n_per, length(idx))

    local <- seq_along(idx)
    lens  <- as.numeric(sf::st_length(L[idx, , drop = FALSE]))
    lens[!is.finite(lens)] <- 0
    sprob <- sum(lens, na.rm = TRUE)
    prob <- if (sprob > 0 && length(lens) == length(local)) lens / sprob else NULL

    pick_local <- sample(local, size = k, replace = draw_with_replacement, prob = prob)
    pick_idx   <- idx[pick_local]

    piece <- L[pick_idx, , drop = FALSE]
    piece$tract_id <- tracts_sf$tract_id[i]
    piece$stratum  <- tracts_sf$stratum[i]
    picks[[i]] <- piece
  }
  dplyr::bind_rows(picks)
}

# --- GENERAL (2023 non-cycle) pool -------------------------------------------
if (exists("general23_n") && inherits(general23_n, "sf") && nrow(general23_n)) {
  noncycle23 <- sf::st_transform(general23_n, crs_work)
} else {
  # Tag-based selection, then filter to line geoms only
  noncycle23 <- l23 |>
    sf::st_transform(crs_work) |>
    pick_noncycle_strict() |>
    sf::st_make_valid()

  keep_lines <- sf::st_geometry_type(noncycle23, by_geometry = TRUE) %in% c("LINESTRING","MULTILINESTRING")
  noncycle23 <- noncycle23[keep_lines, , drop = FALSE]

  if (nrow(noncycle23)) {
    noncycle23 <- suppressWarnings(sf::st_cast(noncycle23, "LINESTRING"))
    noncycle23 <- noncycle23[!sf::st_is_empty(sf::st_geometry(noncycle23)), , drop = FALSE]
  }

  # Spatial fallback if still empty: base roads minus cycle buffer
  if (!nrow(noncycle23)) {
    base_roads <- c("trunk","primary","secondary","tertiary",
                    "unclassified","residential","service",
                    "primary_link","secondary_link","tertiary_link",
                    "living_street","pedestrian")

    base23 <- l23 |>
      sf::st_transform(crs_work) |>
      dplyr::filter(.data$highway %in% base_roads)

    keep_lines <- sf::st_geometry_type(base23, by_geometry = TRUE) %in% c("LINESTRING","MULTILINESTRING")
    base23 <- base23[keep_lines, , drop = FALSE] |>
              sf::st_make_valid() |>
              suppressWarnings(sf::st_cast("LINESTRING"))

    cyc_geom <- NULL
    if (exists("cyc23_n") && inherits(cyc23_n,"sf") && nrow(cyc23_n)) {
      cyc_geom <- sf::st_buffer(sf::st_geometry(cyc23_n), tol_m) |> sf::st_union()
    } else if (exists("cyc23") && inherits(cyc23,"sf") && nrow(cyc23)) {
      cyc_geom <- sf::st_buffer(sf::st_geometry(sf::st_transform(cyc23, crs_work)), tol_m) |> sf::st_union()
    }

    if (!is.null(cyc_geom) && length(cyc_geom) > 0) {
      diff_geom <- sf::st_difference(sf::st_geometry(base23), cyc_geom)
      diff_geom <- sf::st_collection_extract(diff_geom, "LINESTRING", warn = FALSE)
      noncycle23 <- sf::st_sf(geometry = diff_geom, crs = crs_work)
      noncycle23 <- noncycle23[!sf::st_is_empty(sf::st_geometry(noncycle23)), , drop = FALSE]
    } else {
      noncycle23 <- base23
    }
  }
}

# --- STRATIFIED PICKS (per-tract quotas) --------------------------------------
added_by_tr    <- sample_lines_by_tract(added,   tracts, PERTRACT_ADD, replace = FORCE_QUOTA)
removed_by_tr  <- sample_lines_by_tract(removed, tracts, PERTRACT_REM, replace = FORCE_QUOTA)

# GENERAL: restrict to tracts where we sampled ADD (fallback = all tracts)
sampled_ids <- unique(added_by_tr$tract_id)
tracts_for_general <- dplyr::filter(tracts, tract_id %in% sampled_ids)
if (!nrow(tracts_for_general)) tracts_for_general <- tracts

general_by_tr <- sample_lines_by_tract(noncycle23, tracts_for_general, PERTRACT_GEN, replace = FORCE_QUOTA)

# --- Midpoints (1 point per sampled row) + metadata ---------------------------
midpoint_one <- function(x){
  if (!inherits(x,"sf") || !nrow(x)) return(sf::st_sf(id=integer(0), geometry=sf::st_sfc(crs=sf::st_crs(x))))
  g <- sf::st_geometry(x) |> sf::st_make_valid() |> suppressWarnings(sf::st_line_merge())
  sf::st_as_sf(data.frame(id=seq_len(nrow(x))),
               geometry = sf::st_point_on_surface(g),
               crs = sf::st_crs(x))
}

added_pts   <- midpoint_one(added_by_tr)    |> dplyr::mutate(class="ADD",     tract_id=added_by_tr$tract_id,    stratum=added_by_tr$stratum)
removed_pts <- midpoint_one(removed_by_tr)  |> dplyr::mutate(class="REMOVE",  tract_id=removed_by_tr$tract_id,  stratum=removed_by_tr$stratum)
gen_pts     <- midpoint_one(general_by_tr)  |> dplyr::mutate(class="GENERAL", tract_id=general_by_tr$tract_id,  stratum=general_by_tr$stratum)

# --- Tables with lon/lat + placeholders, then bind ---------------------------
add_tbl <- to_lonlat_tbl(added_pts)   |> dplyr::mutate(interval = "2019→2023") |> add_validation_cols() |> coerce_for_bind()
rem_tbl <- to_lonlat_tbl(removed_pts) |> dplyr::mutate(interval = "2019→2023") |> add_validation_cols() |> coerce_for_bind()
gen_tbl <- to_lonlat_tbl(gen_pts)     |> dplyr::mutate(interval = "2019→2023") |> add_validation_cols() |> coerce_for_bind()

samples_tbl <- dplyr::bind_rows(add_tbl, rem_tbl, gen_tbl)

```

```{r}
#| label: tables-export
#| echo: false
#| message: false
#| warning: false
#| eval: true

stopifnot(exists("add_tbl"), exists("rem_tbl"), exists("gen_tbl"))

# Where to save
outdir  <- "outputs"
dir.create(outdir, showWarnings = FALSE, recursive = TRUE)
outfile <- file.path(outdir, paste0(city_tag, "_samples_2019_2023.xlsx"))

library(openxlsx)
library(dplyr)

wb <- createWorkbook()

write_one <- function(df, sheet){
  df2 <- df %>%
    mutate(GSV = NA_character_) %>%
    relocate(GSV, .after = gsv_link)
  addWorksheet(wb, sheet)
  writeData(wb, sheet, df2)
  if (nrow(df2) > 0) {
    link_col <- which(names(df2) == "GSV")
    writeFormula(
      wb, sheet,
      x = paste0('HYPERLINK("', df$gsv_link, '","Open GSV")'),
      startCol = link_col, startRow = 2
    )
  }
  addFilter(wb, sheet, row = 1, cols = 1:ncol(df2))
  setColWidths(wb, sheet, cols = 1:ncol(df2), widths = "auto")
  freezePane(wb, sheet, firstActiveRow = 2, firstActiveCol = 1)
}

write_one(add_tbl, "ADD")
write_one(rem_tbl, "REMOVE")
write_one(gen_tbl, "GENERAL")

saveWorkbook(wb, outfile, overwrite = TRUE)
message("Excel written: ", normalizePath(outfile))
```

```{r}
#| label: validation-map-points
#| echo: false
#| message: false
#| warning: false

library(leaflet)
library(sf)
library(dplyr)

# ---- helper: prep points (WGS84 + lon/lat + popup) --------------------------
prep_pts <- function(x){
  if (!inherits(x,"sf") || !nrow(x)) return(NULL)
  w  <- st_transform(x, 4326)
  xy <- st_coordinates(w)
  good <- is.finite(xy[,1]) & is.finite(xy[,2])
  if (!any(good)) return(NULL)
  w   <- w[good, , drop = FALSE]
  xy  <- xy[good, , drop = FALSE]
  w$lon <- xy[,1]; w$lat <- xy[,2]
  w$gsv <- sprintf("https://www.google.com/maps/@?api=1&map_action=pano&viewpoint=%f,%f", xy[,2], xy[,1])
  w$popup <- paste0(
    "<b>", w$class, "</b>",
    "<br/>tract: ", w$tract_id,
    if ("stratum" %in% names(w)) paste0("<br/>stratum: ", w$stratum) else "",
    "<br/><a href='", w$gsv, "' target='_blank'>Open GSV</a>"
  )
  w
}

add_wgs <- prep_pts(added_pts)
rem_wgs <- prep_pts(removed_pts)
gen_wgs <- prep_pts(gen_pts)

# ---- bounds (prefer city_perimeter -> points -> tracts -> fallback) ----------
if (exists("city_perimeter")) {
  bb <- st_bbox(st_transform(city_perimeter, 4326))
  bounds <- unname(c(bb$xmin, bb$ymin, bb$xmax, bb$ymax))
} else {
  bbs <- list()
  if (!is.null(add_wgs) && nrow(add_wgs)) bbs <- c(bbs, list(st_bbox(add_wgs)))
  if (!is.null(rem_wgs) && nrow(rem_wgs)) bbs <- c(bbs, list(st_bbox(rem_wgs)))
  if (!is.null(gen_wgs) && nrow(gen_wgs)) bbs <- c(bbs, list(st_bbox(gen_wgs)))
  if (!length(bbs) && exists("tracts") && nrow(tracts)) bbs <- list(st_bbox(st_transform(tracts, 4326)))
  bounds <- if (length(bbs)) {
    M <- do.call(rbind, lapply(bbs, function(b) unname(c(b$xmin,b$ymin,b$xmax,b$ymax))))
    c(min(M[,1]), min(M[,2]), max(M[,3]), max(M[,4]))
  } else c(2.05, 41.30, 2.25, 41.45)  # BCN-ish fallback
}

# ---- colours -----------------------------------------------------------------
# points (Dark2-ish, no purple)
col_add_pt <- "#E6AB02"  # gold
col_rem_pt <- "#D95F02"  # orange-red
col_gen_pt <- "#1B9E77"  # teal
stroke     <- "#000000"

# tracts
tract_stroke <- "#000000"   # very light outline
tract_fill   <- "#FFFFFF"   # fully transparent fill


# networks (optional)
col_2019_net   <- "#666666"
col_added_net  <- "#D95F02"
col_removedNet <- "#FFFFFF"
col_removedHalo<- "#000000"
col_2023_net   <- "#1B9E77"
col_nonci      <- "#BDBDBD"   # <-- NEW: general (non-CI) net


# ---- map ---------------------------------------------------------------------
m <- leaflet(options = leafletOptions(preferCanvas = TRUE)) %>%
  addProviderTiles("CartoDB.Positron", group = "Positron")

# Tracts polygons (transparent fill, faint outline)
if (exists("tracts") && inherits(tracts,"sf") && nrow(tracts)) {
  m <- m %>%
    # put tracts below points but above tiles (optional, for layering)
    addMapPane("tractsPane", zIndex = 410) %>%
    addPolygons(
     data = st_transform(tracts, 4326), group = "Tracts",
      color = tract_stroke, opacity = 0.7,  # <-- visible but subtle
      weight = 0.8,                          # hairline; try 1.0 if still faint
      fillColor = tract_fill, fillOpacity = 0,
      smoothFactor = 0.5
  )

  # Area labels (already transparent background)
  tract_centers <- st_point_on_surface(st_transform(tracts, 4326))
  m <- m %>%
    addCircleMarkers(
      data = tract_centers, group = "Area labels",
      radius = 0.1, stroke = FALSE, fillOpacity = 0,
      label = ~as.character(tract_id),
      labelOptions = labelOptions(
        noHide = TRUE, direction = "center",
        style = list(
          "background"      = "transparent",
          "border"          = "none",
          "box-shadow"      = "none",
          "padding"         = "0px",
          "color"           = "rgba(0,0,0,0.25)",
          "text-shadow"     = "none",
          "pointer-events"  = "none"
        )
      )
    )
}

# Prepare General (non-CI) network (lines only)
gnet <- NULL
if (exists("general23_n") && inherits(general23_n,"sf") && nrow(general23_n)) {
  gnet <- general23_n
} else if (exists("noncycle23") && inherits(noncycle23,"sf") && nrow(noncycle23)) {
  gnet <- noncycle23
}

if (!is.null(gnet)) {
  keep_lines <- sf::st_geometry_type(gnet, by_geometry = TRUE) %in% c("LINESTRING","MULTILINESTRING")
  gnet <- gnet[keep_lines, , drop = FALSE]
  if (nrow(gnet)) {
    gnet <- suppressWarnings(sf::st_cast(gnet, "LINESTRING"))
    gnet <- gnet[!sf::st_is_empty(sf::st_geometry(gnet)), , drop = FALSE]
    gnet_wgs <- sf::st_transform(gnet, 4326)
  } else {
    gnet_wgs <- NULL
  }
} else {
  gnet_wgs <- NULL
}

# Optional network context (added but hidden by default below)
if (exists("cyc19_n") && inherits(cyc19_n,"sf") && nrow(cyc19_n))
  m <- m %>% addPolylines(data = st_transform(cyc19_n, 4326), group = "2019 net",
                          weight = 2, color = col_2019_net, opacity = 0.7)

if (exists("removed") && inherits(removed,"sf") && nrow(removed)) {
  m <- m %>% addPolylines(data = st_transform(removed, 4326), group = "Removed net",
                          weight = 2.6, color = col_removedHalo, opacity = 0.18)
  m <- m %>% addPolylines(data = st_transform(removed, 4326), group = "Removed net",
                          weight = 2.0, color = col_removedNet,  opacity = 1.0)
}

if (exists("added") && inherits(added,"sf") && nrow(added))
  m <- m %>% addPolylines(data = st_transform(added, 4326), group = "Added net",
                          weight = 2, color = col_added_net, opacity = 0.95)

if (exists("cyc23_n") && inherits(cyc23_n,"sf") && nrow(cyc23_n))
  m <- m %>% addPolylines(data = st_transform(cyc23_n, 4326), group = "2023 net",
                          weight = 2, color = col_2023_net, opacity = 0.6)

# Points (explicit lng/lat; active by default)
if (!is.null(add_wgs) && nrow(add_wgs) > 0)
  m <- m %>% addCircleMarkers(data = add_wgs, lng = ~lon, lat = ~lat, group = "ADD pts",
                              radius = 4, weight = 1, color = stroke,
                              fillColor = col_add_pt, fillOpacity = 0.95,
                              popup = ~popup)

if (!is.null(rem_wgs) && nrow(rem_wgs) > 0)
  m <- m %>% addCircleMarkers(data = rem_wgs, lng = ~lon, lat = ~lat, group = "REMOVE pts",
                              radius = 4, weight = 1, color = stroke,
                              fillColor = col_rem_pt, fillOpacity = 0.95,
                              popup = ~popup)

if (!is.null(gen_wgs) && nrow(gen_wgs) > 0)
  m <- m %>% addCircleMarkers(data = gen_wgs, lng = ~lon, lat = ~lat, group = "GENERAL pts",
                              radius = 4, weight = 1, color = stroke,
                              fillColor = col_gen_pt, fillOpacity = 0.95,
                              popup = ~popup)
# General (non-CI) network
if (!is.null(gnet_wgs) && nrow(gnet_wgs))
  m <- m %>% addPolylines(data = gnet_wgs, group = "General (non-CI) net",
                          weight = 1, color = col_nonci, opacity = 0.7)

# Control + single legend
m %>%
  clearControls() %>%
  addLayersControl(
    baseGroups = c("Positron"),
    overlayGroups = c(
      "Tracts","Area labels",
      "2019 net","Added net","Removed net","2023 net",
      "General (non-CI) net",         # <-- add here
      "ADD pts","REMOVE pts","GENERAL pts"
    ),
    options = layersControlOptions(collapsed = TRUE)
  ) %>%
  hideGroup(c("2019 net","Added net","Removed net","2023 net","General (non-CI) net")) %>%  # <-- add here
  fitBounds(bounds[1], bounds[2], bounds[3], bounds[4]) %>%
  addLegend(
    position = "bottomright", title = "",
    colors = c(col_add_pt, col_rem_pt, col_gen_pt),
    labels = c("ADD points", "REMOVE points", "GENERAL points"),
    opacity = 1
  )



```

### Outcomes and metrics

-   **Precision (ADD)**:

    -   TP_ADD = points coded as ADD and verifiable in 2023, with infrastructure present.

    -   FP_ADD = points coded as ADD and verifiable in 2023, but infrastructure absent.

    -   Precision_ADD = TP_ADD / (TP_ADD + FP_ADD).

-   **Precision (REMOVE)**:

    -   Defined analogously, with present_2023 = FALSE as the true positive condition.

-   **False negatives (GENERAL)**:

    -   Strict FN: points coded as GENERAL, verifiable in both years, absent in 2019 but present in 2023.

    -   FN_rate = FN_strict / N_strict, where N_strict = number of GENERAL points verifiable in both years.

-   **Recall (ADD)**:

    -   Recall_ADD = TP_ADD / (TP_ADD + FN_strict).

-   **Calibration of OSM km (adds)**:

    -   Adjustment of raw OSM-added kilometres using observed Precision_ADD and Recall_ADD.

$$
K_{\text{add, validated}}
\approx
K_{\text{add, OSM}}
\times
\frac{1 - \mathrm{FP}_{\text{ADD}}}{1 - \mathrm{FN}_{\text{ADD}}}
$$ {#eq-kadd}

(Analogous down-adjustment for removals using FP_REMOVE.)

<!-- Statistical analysis -->

<!-- Report proportions with 95% CIs (e.g., prop.test). -->

<!-- By-stratum estimates (centrality × density); compute a weighted overall (equal weights if equal per stratum; otherwise use design weights reflecting tract/pop coverage). -->

<!-- Sensitivity analyses: -->

<!-- “Loose FN”: verif_2023=TRUE & present_2023=TRUE (regardless of 2019). -->

<!-- Alternative tolerance windows (±0/±1 year). -->

<!-- Inter-rater reliability: double-code ~10% of points; report κ. -->

<!-- Reproducibility and ethics -->

<!-- Provide code (R/Quarto) and exact tag list; no personal data; imagery used is publicly accessible via GSV. -->

<!-- Results (template to fill once coded) -->

<!-- Sample flow. Counts by class and verifiability (table). -->

<!-- ADD: TP/FP counts, Precision_ADD with CI; by-stratum differences (heatmap/table). -->

<!-- REMOVE: TP/FP counts, Precision_REMOVE with CI; by-stratum. -->

<!-- GENERAL: FN_strict and FN_rate with CI; Recall_ADD with CI. -->

<!-- Calibration. Raw OSM km vs validated km (point estimate + CI); % change after adjustment. -->

<!-- Sensitivity. Changes under loose FN and alternative windows. -->

<!-- Reliability. κ for double-coded subset. -->

<!-- Discussion -->

<!-- Principal findings. (e.g., OSM precision high for ADD, moderate for REMOVE; FN_rate non-trivial in peripheral strata; recall X%.) -->

<!-- Implications. When and where OSM is dependable for monitoring CI; how to adjust OSM-based indicators. -->

<!-- Biases. Systematic error by urban context (density/centrality), imagery gaps. -->

<!-- Strengths & limitations. Stratified design; portable workflow; GSV timing constraints; sample size. -->

<!-- Relation to prior work. How estimates compare with municipal reports / prior validations. -->

<!-- Conclusions -->

<!-- OSM is [adequate / limited] for detecting CI changes when [conditions]. -->

<!-- Provide validated growth estimates with uncertainty; recommend routine stratified validation for cross-city comparability. -->

<!-- Tables & Figures (recommended) -->

<!-- T1. Sample counts by class × verifiability (and by stratum). -->

<!-- T2. ADD/REMOVE precision with 95% CIs (overall + by stratum). -->

<!-- T3. GENERAL FN_rate and Recall_ADD with 95% CIs (overall + by stratum). -->

<!-- F1. Stratified sampling map (tracts + sampled points). -->

<!-- F2. Heatmap of metrics by stratum. -->

<!-- F3. Raw vs validated km (adds/removals) with CIs (bar/CI plot). -->

<!-- Supp. Coding protocol examples (what counts as CI), sensitivity plots. -->

<!-- ## References -->
