---
# title: "ATRAPA Work Package 1: Built Environment Transformations Dataset"
format: html
editor: visual
bibliography: wp1.bib
---

## OSM validation method

### Aim

Validate OSM cycling infrastructure in Barcelona across years 2019 and 2023.

### Design

-   Stratified random sample of census tracts (centrality × density, 3×3).

-   Five tracts per stratum (45 in total).

-   Per tract and year:

    -   3 random points on the OSM cycleway network (*if present in tract*) → check for false positives for that year.

    -   3 random points on the OSM general road network → check for false negative for that year.

-   Validation source: Google Street View (GSV).

### OSM data

```{r include=FALSE}
#remotes::install_github("ropensci/osmextract")
library(osmdata)
library(osmextract)
library(sf)
library(tidyverse)
library(tmap)
library(grid)
library(gridExtra)
library(spdep)
```

```{r boundaries}
#| label: boundaries
#| eval: false
#| include: false

bb <- getbb("Barcelona, Spain")

od <- opq(bb) |>
  add_osm_feature("boundary", "administrative") |>
  add_osm_feature("admin_level", "8") |>
  add_osm_feature("name", "Barcelona") |>
  osmdata_sf()

perim <- od$osm_multipolygons
if (is.null(perim) || nrow(perim) == 0) perim <- od$osm_polygons
stopifnot(nrow(perim) > 0)

# Prefer municipality (Wikidata Q1492) else largest polygon
if ("wikidata" %in% names(perim) && any(perim$wikidata == "Q1492")) {
  perim <- perim[perim$wikidata == "Q1492", ]
} else {
  perim <- perim[which.max(st_area(perim)), ]
}

perim <- st_make_valid(perim) |> st_cast("MULTIPOLYGON") |> st_set_crs(4326)

st_write(perim, "data/barcelona_perimeter.gpkg",
         layer = "barcelona_perimeter", driver = "GPKG", append = FALSE)
```

```{r infras}
#| label: infras
#| eval: false
#| include: false
perimeter_path <- "data/barcelona_perimeter.gpkg"
stopifnot(file.exists(perimeter_path))
city_perimeter <- sf::st_read(perimeter_path, layer = "barcelona_perimeter", quiet = TRUE)
if (is.na(sf::st_crs(city_perimeter)$epsg) || sf::st_crs(city_perimeter)$epsg != 4326) {
  city_perimeter <- sf::st_transform(city_perimeter, 4326)
}

allowed_highway <- c(
  "living_street","pedestrian","cycleway","motorway","trunk","primary","secondary",
  "tertiary","unclassified","residential","motorway_link","trunk_link","primary_link",
  "secondary_link","tertiary_link","service","track","bus_guideway","escape","raceway","busway"
)

fetch_and_crop_data <- function(version, perimeter, output_file) {
  if (file.exists(output_file)) {
    message("↪ Skipping (exists): ", basename(output_file)); return(invisible(NULL))
  }
  # Always pull from "spain" (works for 2017/2019/2023), read only 'lines'
  lines_data <- tryCatch({
    oe_get(
      place   = "spain",
      version = version,
      layer   = "lines",
      boundary = st_bbox(perimeter), boundary_type = "clipsrc",  # fast pre-clip
      extra_tags = c("cycleway","cycleway:left","cycleway:right","bicycle","route"),
      quiet = FALSE
    )
  }, error = function(e) { message("Error fetching ", version, ": ", e$message); return(NULL) })

  if (is.null(lines_data) || nrow(lines_data) == 0) {
    message("No data returned for version ", version); return(invisible(NULL))
  }

  # Keep only (multi)lines
  is_line <- sf::st_geometry_type(lines_data) %in% c("LINESTRING","MULTILINESTRING")
  lines_data <- lines_data[is_line, ]
  if (nrow(lines_data) == 0) { message("No LINESTRINGs for version ", version); return(invisible(NULL)) }

  # Filter to common road classes (do it in R to avoid GDAL vectortranslate issues)
  if ("highway" %in% names(lines_data)) {
    lines_data <- dplyr::filter(lines_data, highway %in% allowed_highway)
  }

  # Final polygon clip
  perim_valid <- sf::st_make_valid(perimeter)
  lines_data <- lines_data[sf::st_intersects(lines_data, perim_valid, sparse = FALSE), ]
  lines_data <- sf::st_intersection(lines_data, perim_valid)

  out_layer <- sub("\\.gpkg$", "", basename(output_file))
  sf::st_write(lines_data, output_file, layer = out_layer, driver = "GPKG", append = FALSE, quiet = TRUE)
  message("✓ Saved ", basename(output_file), " (", nrow(lines_data), " features)")
}

# Exactly these years (add 2017 if you want by uncommenting)
# fetch_and_crop_data("170101", city_perimeter, "data/barcelona_170101_lines.gpkg")
fetch_and_crop_data("200101", city_perimeter, "data/barcelona_19_lines.gpkg")
fetch_and_crop_data("240101", city_perimeter, "data/barcelona_23_lines.gpkg")

message("Done: Barcelona (files in data/)")
```

```{r}
#| label: infraplot
#| eval: false
#| include: false
# --- cities to plot (match your file prefixes under data/) ---
city_tags <- c("barcelona")      # we saved data/barcelona_*.gpkg
versions  <- c("19","23")  # 2019 & 2023

# --- palette & helpers ---
pal <- c("cycleway"="#E41A1C","pedestrian"="#377EB8","living street"="#4DAF4A","others"="grey")

read_perimeter <- function(tag){
  perim_path <- file.path("data", paste0(tag, "_perimeter.gpkg"))  # << no subfolder
  if (!file.exists(perim_path)) return(NULL)
  # read specific layer and ensure WGS84
  sf::st_read(perim_path, layer = paste0(tag, "_perimeter"), quiet = TRUE) |>
    sf::st_transform(4326)
}

read_infra <- function(tag, ver){
  gpkg <- file.path("data", paste0(tag, "_", ver, "_lines.gpkg"))  # << no subfolder
  lyr  <- paste0(tag, "_", ver, "_lines")
  if (!file.exists(gpkg)) return(NULL)
  x <- tryCatch(sf::st_read(gpkg, layer = lyr, quiet = TRUE), error = function(e) NULL)
  x
}

mk_aggr <- function(x, perimeter){
  if (is.null(x) || nrow(x) == 0) return(NULL)
  x <- sf::st_transform(x, sf::st_crs(perimeter))
  if (!"highway_aggr" %in% names(x)) {
    x <- x |>
      dplyr::mutate(highway_aggr = dplyr::case_when(
        highway == "cycleway"      ~ "cycleway",
        highway == "pedestrian"    ~ "pedestrian",
        highway == "living_street" ~ "living street",
        TRUE ~ "others"
      ))
  }
  x
}

mk_map <- function(perimeter, x, title){
  if (is.null(x) || nrow(x) == 0) {
    return(tm_shape(perimeter) + tm_borders() + tm_title(paste0(title, " (no data)")))
  }
  tm_shape(perimeter) + tm_borders() +
    tm_shape(x) +
    tm_lines(col = "highway_aggr", lwd = 0.5, col_alpha = 0.6,
             col.scale = tm_scale(values = pal)) +
    tm_title(title)
}

make_city_panel <- function(city_tag){
  perim <- read_perimeter(city_tag)
  if (is.null(perim)) return(NULL)

  x19 <- mk_aggr(read_infra(city_tag, "19"), perim)
  x23 <- mk_aggr(read_infra(city_tag, "23"), perim)

  m19 <- mk_map(perim, x19, "2019")   # << fixed var name
  m23 <- mk_map(perim, x23, "2023")   # << fixed var name
  tmap_arrange(m19, m23, nrow = 1)
}

# --- display all city panels ---
tmap_mode("view")
panels <- lapply(city_tags, make_city_panel)
invisible(lapply(panels[!sapply(panels, is.null)], print))
```

### Tract data

```{r}
#| include: false
# ================================================================
# STRATIFIED SAMPLING AND STREET VIEW LINK GENERATION
# ================================================================

# ---- Load required libraries ----
library(sf)
library(dplyr)
library(tidyr)
library(biscale)
library(ggplot2)
library(cowplot)
library(osmdata)
library(purrr)
library(jsonlite)
library(httr)
library(tmap)
library(stringr)
library(sf)
library(leaflet)
library(htmltools)
```

```{r}
#| include: false
# ================================================================
# 1. LOAD AND PREPARE DATA
# ================================================================

# Load spatial and population data
tracts_geo <- readRDS("data/bcn-seccio-censal/BCN_seccion_censal_2015.rds")
tracts_pop <- readRDS("data/pop/Population_Census_tract_2015_2022.rds")

tracts_pop_wide <- tracts_pop |> 
  filter(Year %in% c("2015", "2019", "2022")) |> 
  pivot_wider(names_from = Year, values_from = Population, names_prefix = "pop_")

# Merge geometries with population data
barcelona_tracts <- left_join(tracts_geo, tracts_pop_wide, by = c("CUSEC" = "GEOID")) |> 
  st_transform(25831)

# Compute area, population density, and distance to city center
placa_catalunya <- st_sfc(st_point(c(431870, 4581450)), crs = 25831)

barcelona_tracts <- barcelona_tracts |> 
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6,
    dens_2015 = pop_2015 / area_km2,
    dens_2019 = pop_2019 / area_km2,
    dens_2022 = pop_2022 / area_km2,
    centroid = st_centroid(geometry),
    dist_centre_km = as.numeric(st_distance(centroid, placa_catalunya)) / 1000
  )
```

```{r}
#| message: false
#| warning: false
#| include: false
#| paged-print: false

# ================================================================
# 2. STRATIFIED SAMPLING
# ================================================================

barcelona_tracts <- barcelona_tracts |>
  mutate(
    dens_stratum = ntile(dens_2022, 3),
    cent_stratum = ntile(-dist_centre_km, 3),
    stratum_id = paste0("D", dens_stratum, "_C", cent_stratum)
  )

set.seed(123)
per_stratum <- 5  # try 3–6 depending on effort

sampled_tracts <- barcelona_tracts |> 
  group_by(stratum_id) |> 
  sample_n(size = min(per_stratum, n())) |> 
  ungroup()
```

```{r}
#| echo: false

# ================================================================
# 3. BIVARIATE MAPPING
# ================================================================

barcelona_tracts <- barcelona_tracts |> 
mutate(centrality_flipped = -dist_centre_km)
 
bb_bivar <- bi_class(barcelona_tracts, x = dens_2022, y = centrality_flipped, style = "quantile", dim = 3)
 
bivar_map <- ggplot() +
   geom_sf(data = bb_bivar, aes(fill = bi_class), color = "white", size = 0.1, show.legend = FALSE) +
   bi_scale_fill(pal = "DkBlue2", dim = 3) +
   geom_sf(data = sampled_tracts, fill = NA, color = "black", linewidth = 0.5) +
   labs(title = "") +
   bi_theme() +
   theme(plot.title = element_text(size = 12, hjust = 0.5))
 
 bivar_legend <- bi_legend(pal = "DkBlue2", dim = 3, xlab = "Higher Density", ylab = "More Central", size = 9) +
   theme(axis.title = element_text(size = 9), axis.text = element_blank(),
         plot.margin = margin(5, 5, 5, 5), panel.background = element_blank(), plot.background = element_blank())
 
 final_plot <- ggdraw() +
   draw_plot(bivar_map, 0, 0, 1, 1) +
   draw_plot(bivar_legend, x = 0.73, y = 0.03, width = 0.26, height = 0.26)
 
 final_plot
 
 ggsave("figs/stratified_sample_bivariate_map.png", final_plot, width = 8, height = 8, dpi = 300)
```

### GSV data

```{r}
# =========================================================
# BCN OSM validation — cycle + general (links-only)
# target_year + allowed_from / allowed_to (month-only)
# Paired 2019 & 2023
# =========================================================
library(sf); library(dplyr); library(readr); library(purrr); library(tidyr)
library(tibble); library(stringr); library(lubridate)

set.seed(42)

# ---------------- CONFIG ----------------
CITY               <- "barcelona"
GPKG19             <- file.path("data", paste0(CITY, "_19_lines.gpkg"))
GPKG23             <- file.path("data", paste0(CITY, "_23_lines.gpkg"))
LAYER19            <- paste0(CITY, "_19_lines")
LAYER23            <- paste0(CITY, "_23_lines")

TRACT_ID_COL       <- "CUSEC"     # id column in barcelona_tracts
TRACTS_PER_STRATUM <- 1           # ≥1 tract per (density × centrality) stratum

# quotas per tract × target_year × class
N_PER_TRACT_YEAR_CY <- 2          # cycle points
N_PER_TRACT_YEAR_GN <- 2          # general (non-cycle) points

# sampling knobs
OVERSAMPLE <- 30                  # candidate lines tried before thinning
MIN_DIST_M <- 30                  # min spacing between sampled points (meters)

CRS_METRIC  <- 25831              # ETRS89 / UTM 31N (Barcelona)

# outputs
OUT_DIR          <- "outputs"
OUT_CSV_COMBINED <- file.path(OUT_DIR, paste0(CITY, "_pilot_links_combined.csv"))
OUT_CSV_CYCLE    <- file.path(OUT_DIR, paste0(CITY, "_pilot_links_cycle.csv"))
OUT_CSV_GENERAL  <- file.path(OUT_DIR, paste0(CITY, "_pilot_links_general.csv"))
OUT_GPKG         <- file.path(OUT_DIR, paste0(CITY, "_pilot_points.gpkg"))
MAKE_XLSX        <- TRUE
XLSX_PATH        <- file.path(OUT_DIR, paste0(CITY, "_validation_links.xlsx"))

# ---------------- HELPERS ----------------
`%||%` <- function(a,b) if (is.null(a)) b else a

read_infra <- function(path, layer){
  stopifnot(file.exists(path))
  suppressWarnings(st_read(path, layer = layer, quiet = TRUE)) |> st_make_valid()
}

flag_cycle <- function(x){
  low <- function(n) if (n %in% names(x)) tolower(as.character(x[[n]])) else rep(NA_character_, nrow(x))
  hw  <- low("highway")
  cy  <- low("cycleway"); cyl <- low("cycleway:left"); cyr <- low("cycleway:right")
  ( !is.na(hw) & hw == "cycleway" ) |
  ( !is.na(cy)  & cy  %in% c("lane","track","opposite","opposite_lane","opposite_track","shared_lane","shared_busway","busway") ) |
  ( !is.na(cyl) & cyl %in% c("lane","track","opposite","opposite_lane","opposite_track") ) |
  ( !is.na(cyr) & cyr %in% c("lane","track","opposite","opposite_lane","opposite_track") )
}

stratum_labels <- function(v){
  qs <- stats::quantile(v, probs = c(0, 1/3, 2/3, 1), na.rm = TRUE)
  cut(v, qs, include.lowest = TRUE, labels = c("low","mid","high"))
}

thin_pts <- function(pts, min_dist_m){
  if (nrow(pts) <= 1) return(pts)
  ptsm <- sf::st_transform(pts, CRS_METRIC)
  keep <- logical(nrow(ptsm))
  for (i in seq_len(nrow(ptsm))){
    if (!keep[i]){
      keep[i] <- TRUE
      d <- as.numeric(sf::st_distance(ptsm[i,], ptsm))
      too_close <- setdiff(which(d < min_dist_m), i)
      if (length(too_close)) keep[too_close] <- FALSE
    }
  }
  pts[keep,]
}

gsv_link <- function(pt){
  xy <- sf::st_coordinates(pt) # lon, lat
  paste0("https://www.google.com/maps/@?api=1&map_action=pano&source=outdoor&viewpoint=",
         xy[2], ",", xy[1])
}

# robust union sampler (mixed geoms OK; returns WGS84 points)
sample_union_points <- function(lines19, lines23, area, keep_n, over_n, min_dist_m){
  l19 <- sf::st_transform(lines19, CRS_METRIC)
  l23 <- sf::st_transform(lines23, CRS_METRIC)
  a   <- sf::st_transform(area,    CRS_METRIC)

  idx19 <- lengths(sf::st_intersects(l19, a)) > 0
  idx23 <- lengths(sf::st_intersects(l23, a)) > 0
  if (!any(idx19) && !any(idx23)) return(sf::st_sf(geometry = sf::st_sfc(), crs = 4326))
  c19 <- if (any(idx19)) suppressWarnings(sf::st_intersection(l19[idx19,], a)) else l19[0,]
  c23 <- if (any(idx23)) suppressWarnings(sf::st_intersection(l23[idx23,], a)) else l23[0,]

  sfc_all <- sf::st_sfc(c(sf::st_geometry(c19), sf::st_geometry(c23)), crs = CRS_METRIC)
  ls <- suppressWarnings(sf::st_cast(sfc_all, "LINESTRING", warn = FALSE))
  if (length(ls) == 0) return(sf::st_sf(geometry = sf::st_sfc(), crs = 4326))

  has_len <- as.numeric(sf::st_length(ls)) > 0
  ls <- ls[has_len]
  if (length(ls) == 0) return(sf::st_sf(geometry = sf::st_sfc(), crs = 4326))

  sel_idx <- sample.int(length(ls), size = min(over_n, length(ls)), replace = FALSE)
  ls_sel  <- ls[sel_idx]

  pts_metric <- sf::st_line_sample(ls_sel, n = rep(1, length(ls_sel))) |> sf::st_cast("POINT")
  if (length(pts_metric) == 0) return(sf::st_sf(geometry = sf::st_sfc(), crs = 4326))

  pts_sf <- sf::st_sf(geometry = pts_metric, crs = CRS_METRIC)
  pts_sf <- thin_pts(pts_sf, min_dist_m)
  # after: pts_sf <- thin_pts(pts_sf, min_dist_m)
if (nrow(pts_sf) > keep_n) pts_sf <- pts_sf[seq_len(keep_n), ]  # ← hard cap per tract × class

  sf::st_transform(pts_sf, 4326)
}

# ---------- rows: allowed_from (prev June) + allowed_to (next May) ----------
mk_year_rows <- function(pts, target_year, dens_cat, cent_cat, tract_id, class_label, city = CITY){
  start <- ymd(sprintf("%d-06-01", as.integer(target_year) - 1L))      # e.g., 2019 -> 2018-06-01
  end   <- (start %m+% months(12)) - days(1)                           # -> 2019-05-31

  tibble(
    city         = city,
    target_year  = target_year,          # single source of truth
    osm_class    = class_label,          # "cycle" | "general"
    dens_cat     = dens_cat,
    cent_cat     = cent_cat,
    tract_id     = tract_id,
    lon          = sf::st_coordinates(pts)[,1],
    lat          = sf::st_coordinates(pts)[,2],
    gsv_link     = purrr::map_chr(sf::st_geometry(pts), gsv_link),
    allowed_from = format(start, "%Y-%m"),   # month-only
    allowed_to   = format(end,   "%Y-%m"),   # month-only
    # rater-filled:
    pano_year    = NA_integer_,
    pano_month   = NA_integer_,
    pano_within_window = NA,             # TRUE/FALSE
    label_status = NA_character_,        # OK / NO_PANO / NO_YEAR / MOVE_TOO_FAR / OCCLUDED
    infra_present= NA_character_,        # YES / NO / UNCERTAIN
    infra_type   = NA_character_,        # painted / protected / bus+bike / other
    position_ok  = NA_character_,        # YES / NO
    confidence   = NA_integer_,
    notes        = NA_character_
  )
}

# ---------------- DATA ----------------
stopifnot(exists("barcelona_tracts"), inherits(barcelona_tracts, "sf"))
stopifnot(all(c(TRACT_ID_COL, "dens_2022", "dist_centre_km") %in% names(barcelona_tracts)))

tracts <- barcelona_tracts |>
  dplyr::select(!!TRACT_ID_COL, dens_2022, dist_centre_km, geometry) |>
  dplyr::rename(tract_id = !!rlang::sym(TRACT_ID_COL)) |>
  dplyr::mutate(
    dens_cat = stratum_labels(dens_2022),
    cent_cat = stratum_labels(-dist_centre_km)  # closer = more central
  )

ln19_all <- read_infra(GPKG19, LAYER19)
ln23_all <- read_infra(GPKG23, LAYER23)

is_cyc_19 <- flag_cycle(ln19_all)
is_cyc_23 <- flag_cycle(ln23_all)

cyc19 <- ln19_all[ is_cyc_19, ]
cyc23 <- ln23_all[ is_cyc_23, ]
gen19 <- ln19_all[!is_cyc_19, ]
gen23 <- ln23_all[!is_cyc_23, ]

# choose ≥1 tract per stratum (prioritize cycle length in BOTH years)
tracts_m <- sf::st_transform(tracts, CRS_METRIC)
cyc19_m  <- sf::st_transform(cyc19, CRS_METRIC)
cyc23_m  <- sf::st_transform(cyc23, CRS_METRIC)

len19 <- suppressWarnings(sf::st_intersection(tracts_m, cyc19_m)) |>
  dplyr::mutate(len_m = as.numeric(sf::st_length(geometry))) |>
  sf::st_drop_geometry() |>
  dplyr::group_by(tract_id) |>
  dplyr::summarise(len_m_2019 = sum(len_m, na.rm = TRUE), .groups = "drop")

len23 <- suppressWarnings(sf::st_intersection(tracts_m, cyc23_m)) |>
  dplyr::mutate(len_m = as.numeric(sf::st_length(geometry))) |>
  sf::st_drop_geometry() |>
  dplyr::group_by(tract_id) |>
  dplyr::summarise(len_m_2023 = sum(len_m, na.rm = TRUE), .groups = "drop")

len_tbl <- tracts |>
  sf::st_drop_geometry() |>
  dplyr::select(tract_id, dens_cat, cent_cat) |>
  dplyr::left_join(len19, by = "tract_id") |>
  dplyr::left_join(len23, by = "tract_id") |>
  dplyr::mutate(
    len_m_2019 = len_m_2019 %||% 0,
    len_m_2023 = len_m_2023 %||% 0,
    min_len    = pmin(len_m_2019, len_m_2023, na.rm = TRUE),
    tot_len    = len_m_2019 + len_m_2023
  )

chosen_tbl <- len_tbl %>%
  dplyr::group_by(dens_cat, cent_cat) %>%
  dplyr::arrange(dplyr::desc(min_len), dplyr::desc(tot_len)) %>%
  dplyr::slice_head(n = TRACTS_PER_STRATUM) %>%   # ← uses the knob
  dplyr::ungroup()

chosen_sf <- tracts |> dplyr::inner_join(chosen_tbl, by = c("tract_id","dens_cat","cent_cat"))

# ---------------- SAMPLE & BUILD ROWS ----------------
collect_pts <- function(tr_row){
  out <- list()

  # CYCLE class (paired 2019 & 2023)
  if ((tr_row$len_m_2019 + tr_row$len_m_2023) > 0){
    pts_cy <- sample_union_points(cyc19, cyc23, tr_row, N_PER_TRACT_YEAR_CY, OVERSAMPLE, MIN_DIST_M)
    if (nrow(pts_cy)){
      out <- append(out, list(
        mk_year_rows(pts_cy, 2019L, tr_row$dens_cat, tr_row$cent_cat, tr_row$tract_id, "cycle"),
        mk_year_rows(pts_cy, 2023L, tr_row$dens_cat, tr_row$cent_cat, tr_row$tract_id, "cycle")
      ))
    }
  }

  # GENERAL class (paired 2019 & 2023)
  pts_gn <- sample_union_points(gen19, gen23, tr_row, N_PER_TRACT_YEAR_GN, OVERSAMPLE, MIN_DIST_M)
  if (nrow(pts_gn)){
    out <- append(out, list(
      mk_year_rows(pts_gn, 2019L, tr_row$dens_cat, tr_row$cent_cat, tr_row$tract_id, "general"),
      mk_year_rows(pts_gn, 2023L, tr_row$dens_cat, tr_row$cent_cat, tr_row$tract_id, "general")
    ))
  }

  if (length(out)) dplyr::bind_rows(out) else NULL
}

pts_list <- purrr::map(seq_len(nrow(chosen_sf)), ~collect_pts(chosen_sf[.x,]))
pts_list <- pts_list[!vapply(pts_list, is.null, logical(1))]
res <- if (length(pts_list)) dplyr::bind_rows(pts_list) else tibble()

# ---------------- IDs, compute within-window, EXPORTS ----------------
dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)

if (nrow(res)){
  # keep window columns right after the link
  res <- res %>% dplyr::relocate(allowed_from, allowed_to, .after = gsv_link)

  # compute within-window (>= allowed_from, < allowed_from + 12 months) without storing helper dates
  af <- ymd(paste0(res$allowed_from, "-01"))
  pd <- make_date(res$pano_year, res$pano_month, 1)
  res$pano_within_window <- !is.na(pd) & pd >= af & pd < (af %m+% months(12))

  # IDs and final schema
  res <- res |>
    dplyr::arrange(osm_class, dens_cat, cent_cat, tract_id, target_year) |>
    dplyr::group_by(osm_class, dens_cat, cent_cat, tract_id) |>
    dplyr::mutate(point_local_id = dplyr::row_number()) |>
    dplyr::ungroup() |>
    dplyr::mutate(point_id = paste0(
      "BCN_", ifelse(osm_class=="cycle","CY","GN"), "_",
      "D", substr(dens_cat,1,1), "_C", substr(cent_cat,1,1),
      "_T", tract_id, "_", point_local_id
    )) |>
    dplyr::relocate(point_id, .after = target_year) |>
    dplyr::select(
      city, target_year, point_id, osm_class, dens_cat, cent_cat, tract_id,
      lon, lat, gsv_link, allowed_from, allowed_to,
      pano_year, pano_month, pano_within_window, label_status,
      infra_present, infra_type, position_ok, confidence, notes
    )

  # --- FORCE the guidance window columns (month-only) ---
# --- FORCE month-only guidance window (Jun → May) ---
stopifnot("target_year" %in% names(res))

res <- res %>%
  mutate(
    target_year  = as.integer(target_year),
    allowed_from = sprintf("%04d-06", target_year - 1L),  # previous June
    allowed_to   = sprintf("%04d-05", target_year),       # next May
    # compute pano_within_window using YYYY-MM string comparison (no day math)
    pano_ym      = ifelse(is.na(pano_year) | is.na(pano_month),
                          NA_character_,
                          sprintf("%04d-%02d", as.integer(pano_year), as.integer(pano_month))),
    pano_within_window = !is.na(pano_ym) &
                         pano_ym >= allowed_from &
                         pano_ym <= allowed_to
  ) %>%
  # drop any duplicate/legacy columns that might confuse things
  select(-matches("^window_"),
         -matches("^allowed_from\\.[0-9]+$"),
         -matches("^allowed_to\\.[0-9]+$")) %>%
  # put them next to the link
  relocate(allowed_from, allowed_to, .after = gsv_link)


  
  readr::write_csv(res, OUT_CSV_COMBINED)
  readr::write_csv(dplyr::filter(res, osm_class=="cycle"),   OUT_CSV_CYCLE)
  readr::write_csv(dplyr::filter(res, osm_class=="general"), OUT_CSV_GENERAL)
  message("Wrote: ", OUT_CSV_COMBINED)
  message("Also wrote: ", OUT_CSV_CYCLE, " and ", OUT_CSV_GENERAL)

  # GPKG: rebuild to avoid schema mismatch
  if (file.exists(OUT_GPKG)) unlink(OUT_GPKG)
  pts_sf <- sf::st_as_sf(res, coords = c("lon","lat"), crs = 4326, remove = FALSE)
  pts_sf$pano_within_window <- as.integer(pts_sf$pano_within_window)
  pts_sf$confidence <- as.integer(pts_sf$confidence)

  suppressWarnings(st_write(dplyr::filter(pts_sf, osm_class=="cycle",  target_year==2019),
                            OUT_GPKG, layer="points_cycle_2019",   driver="GPKG", append=FALSE, quiet=TRUE))
  suppressWarnings(st_write(dplyr::filter(pts_sf, osm_class=="cycle",  target_year==2023),
                            OUT_GPKG, layer="points_cycle_2023",   driver="GPKG", append= TRUE, quiet=TRUE))
  suppressWarnings(st_write(dplyr::filter(pts_sf, osm_class=="general",target_year==2019),
                            OUT_GPKG, layer="points_general_2019", driver="GPKG", append= TRUE, quiet=TRUE))
  suppressWarnings(st_write(dplyr::filter(pts_sf, osm_class=="general",target_year==2023),
                            OUT_GPKG, layer="points_general_2023", driver="GPKG", append= TRUE, quiet=TRUE))

  # quick summaries
  print(res %>% count(osm_class, target_year, name="n_points"))
  print(res %>% count(osm_class, dens_cat, cent_cat, target_year, name="n_points") %>% arrange(osm_class, dens_cat, cent_cat, target_year))
} else {
  message("No points produced — try raising OVERSAMPLE or lowering MIN_DIST_M.")
}

# ---------------- Optional: Excel with clickable links ----------------
# ---------------- Excel with clickable links (safe) ----------------
if (MAKE_XLSX && nrow(res)){
  if (!requireNamespace("openxlsx", quietly = TRUE)) install.packages("openxlsx")
  library(openxlsx)

  wb <- createWorkbook()
  addWorksheet(wb, "rating")

  # 1) Write the data as-is
  writeData(wb, "rating", res, startRow = 1, startCol = 1, withFilter = TRUE)

  # 2) Force month strings to stay as text (avoid Excel reformat)
  textStyle <- createStyle(numFmt = "@")
  rows <- 2:(nrow(res) + 1)
  col_from <- which(names(res) == "allowed_from")
  col_to   <- which(names(res) == "allowed_to")
  if (length(col_from)) addStyle(wb, "rating", textStyle, rows = rows, cols = col_from, gridExpand = TRUE)
  if (length(col_to))   addStyle(wb, "rating", textStyle, rows = rows, cols = col_to,   gridExpand = TRUE)

  # 3) Append the clickable "Open_GSV" at the END so we don’t overwrite allowed_*
  col_open <- ncol(res) + 1
  writeData(wb, "rating", "Open_GSV", startRow = 1, startCol = col_open)
  formulas <- paste0('HYPERLINK("', res$gsv_link, '","Open GSV")')
  writeFormula(wb, "rating", x = formulas, startRow = 2, startCol = col_open)

  freezePane(wb, "rating", firstRow = TRUE)
  setColWidths(wb, "rating", cols = 1:(ncol(res) + 1), widths = "auto")
  saveWorkbook(wb, XLSX_PATH, overwrite = TRUE)
  message("Excel written: ", XLSX_PATH)
}
```

<!-- ## References -->
